---
title: "EMODnet Biology Geospatial R Tutorials"
---

## Welcome

Learn how to access and analyze marine biological and environmental data from [EMODnet](https://emodnet.ec.europa.eu/en/biology) using R. These hands-on tutorials guide you through real-world marine spatial analysis workflows, from accessing data via web services to creating publication-ready visualizations.

Whether you're investigating marine protected areas, exploring species distributions, or analyzing biodiversity change, these tutorials provide reproducible code and practical examples you can adapt to your own research questions.

::: {.callout-tip}
## New to geospatial analysis?

If you're unfamiliar with spatial data concepts, coordinate systems, or the difference between vector and raster data, start with our [**Geospatial Concepts**](concepts.qmd) page.
:::

## What You'll Learn

These tutorials demonstrate how to:

-   **Access marine data programmatically** using EMODnet's web services (WFS and WCS)
-   **Work with vector data** representing species occurrences, habitat boundaries, and human activities
-   **Analyze raster data** including bathymetry, species distributions, and environmental variables
-   **Integrate multiple data sources** from EMODnet, Copernicus Marine Service, and trait databases
-   **Apply spatial operations** such as intersection, masking, and point extraction
-   **Visualize spatial patterns** relevant to marine conservation and management

By the end of these tutorials, you'll be able to independently access, analyze, and visualize EMODnet data for your own marine research questions.

## Tutorial Series

```{r}
#| echo: false
#| output: asis

library(rmarkdown)

# Get tutorial files in order
tutorial_files <- list.files(
  "tutorials",
  pattern = "^tutorial-\\d+\\.qmd$",
  full.names = TRUE
)
tutorial_files <- tutorial_files[order(tutorial_files)]

# Start grid container
cat('::: {.grid}\n\n')

# Loop through tutorials and render cards
for (tutorial_file in tutorial_files) {
  # Read YAML frontmatter
  yaml_content <- rmarkdown::yaml_front_matter(tutorial_file)

  # Extract tutorial number from title
  tutorial_number <- sub(":.*", "", yaml_content$title)
  short_title <- sub("^\\d+:\\s*", "", yaml_content$title)

  # Prepare parameters for partial
  params <- list(
    number = tutorial_number,
    short_title = short_title,
    subtitle = yaml_content$subtitle,
    difficulty = yaml_content$difficulty,
    time = yaml_content$time,
    data_type = yaml_content$`data-type`,
    description = yaml_content$description,
    datasets = yaml_content$datasets,
    path = tutorial_file
  )

  # Render the partial
  result <- knitr::knit_child(
    input = "_tutorial-card.qmd",
    envir = environment(),
    quiet = TRUE
  )

  cat(result, sep = '\n')
}

# Close grid container
cat('\n:::\n')
```

## Who These Tutorials Are For

These tutorials are designed for:

-   **Marine researchers** working with spatial data on species distributions, habitats, or human activities
-   **Conservation practitioners** conducting marine spatial planning, MPA assessments, or environmental impact analyses
-   **Data scientists** interested in applying geospatial methods to marine applications
-   **Graduate students** learning marine ecology, spatial analysis, or data science
-   **Policy analysts** requiring evidence-based spatial information for marine management

## Prerequisites

To get the most from these tutorials, you should have:

**R Programming:**

-   Basic to intermediate R skills (working with data frames, installing packages, running functions)
-   Familiarity with the tidyverse is helpful but not required
-   Experience with RStudio or similar R development environment

**Geospatial Background:**

-   Understanding of basic spatial concepts (coordinates, projections, spatial data types)
-   Awareness of vector vs. raster data
-   Familiarity with common spatial operations

::: {.callout-note}
If you're new to geospatial analysis, visit our [**Geospatial Concepts**](concepts.qmd) page for an introduction to key concepts before starting the tutorials.
:::

## Getting Started

### Installation

Install the required packages using the following code:

```{r}
#| eval: false
#| code-fold: false

# Install packages from CRAN
install.packages(c(
    "emodnet.wfs",
    "sf",
    "terra",
    "dplyr",
    "ggplot2",
    "tmap"
))

# Install emodnet.wcs from GitHub (not yet on CRAN)
# install.packages("pak")
pak::pak("EMODnet/emodnet.wcs")
```

### Quick Start

1.  **New to geospatial analysis?** → Start with [Geospatial Concepts](concepts.qmd)
2.  **Ready to dive in?** → Begin with [Tutorial 1](tutorials/tutorial-01.qmd) to learn WFS basics
3.  **Interested in raster analysis?** → Jump to [Tutorial 2](tutorials/tutorial-02.qmd) for WCS workflows
4.  **Want to combine data sources?** → Try [Tutorial 3](tutorials/tutorial-03.qmd) or [Tutorial 4](tutorials/tutorial-04.qmd)

## Resources and Links

### EMODnet Resources

-   [EMODnet Biology Portal](https://emodnet.ec.europa.eu/en/biology) - Access marine biological data
-   [EMODnet Seabed Habitats](https://emodnet.ec.europa.eu/en/seabed-habitats) - Habitat maps and classifications
-   [EMODnet Bathymetry](https://emodnet.ec.europa.eu/en/bathymetry) - Depth data and terrain models
-   [EMODnet Human Activities](https://emodnet.ec.europa.eu/en/human-activities) - Infrastructure and activity data

### R Package Documentation

-   [emodnet.wfs on GitHub](https://github.com/EMODnet/emodnet.wfs) - Access WFS vector data
-   [emodnet.wcs on GitHub](https://github.com/EMODnet/emodnet.wcs) - Access WCS raster data
-   [sf package](https://r-spatial.github.io/sf/) - Spatial vector data in R
-   [terra package](https://rspatial.github.io/terra/) - Spatial raster data in R

### Learning Resources

-   [Geocomputation with R](https://r.geocompx.org/) - Comprehensive guide to spatial analysis in R
-   [Spatial Data Science](https://r-spatial.org/book/) - Spatial data methods and applications
-   [sf package vignettes](https://r-spatial.github.io/sf/articles/) - Detailed sf documentation

## About This Project

These tutorials are part of the [EMODnet Biology](https://emodnet.ec.europa.eu/en/biology) initiative. Learn more about the project, contributors, and funding on the [About](about.qmd) page.
