---
title: "4: Functional Traits Across Seabed Habitats"
subtitle: "Exploring Functional Traits Across Seabed Habitats in the Aegean Sea"
difficulty: "Advanced"
time: "~90 minutes"
data-type: "WFS + WCS + External sources"
description: "Integrate EMODnet data with external trait databases and Copernicus Marine Service environmental data. Investigate how benthic species body size varies across habitat types and environmental gradients in the Aegean Sea."
datasets: "Benthic species occurrences, seabed habitat maps, trait databases, temperature reanalysis data"
---

## Tutorial Content

This tutorial is currently under development.

### What You'll Learn

- Demonstrating combined use of emodnet.wfs, emodnet.wcs, and CopernicusMarine packages
- Retrieving and visualising body size traits for benthic species using trait databases
- Exploring trait–environment relationships across habitat types
- Practical geospatial workflows including raster extraction and spatial joins

### Data Sources

- **EMODnet WFS:**
  - `eb185_benthos_full_matched`: Benthic species records in the Aegean
  - `eu_seabed_habitats_2019`: Seabed habitat polygons
- **EMODnet WCS:**
  - `emodnet__mean_2020`: Bathymetry raster (mean depth of seafloor)
- **Copernicus Marine Service:**
  - `GLOBAL_MULTIYEAR_PHY_001_030`: Bottom temperature and/or salinity reanalysis data
- **External Resources:**
  - **MOBS (Marine Organism Body Size) database**: [github.com/crmcclain/MOBS_OPEN](https://github.com/crmcclain/MOBS_OPEN)

### Study Area

Aegean Sea

Check back soon for the complete tutorial!

## CI Debug: Copernicus Marine Download Tests

```{r}
library(CopernicusMarine)
cat(
  "CopernicusMarine version:",
  as.character(packageVersion("CopernicusMarine")),
  "\n"
)
cat("sf version:", as.character(packageVersion("sf")), "\n")
cat("R version:", R.version.string, "\n")
cat("Timezone:", Sys.timezone(), "\n")
cat("Locale:", Sys.getlocale("LC_TIME"), "\n")
cat("blosc support:", CopernicusMarine:::has_blosc, "\n")
cat("lubridate version:", as.character(packageVersion("lubridate")), "\n")
cat("zoneinfo exists:", file.exists("/usr/share/zoneinfo"), "\n")
cat("System:", Sys.info()[c("sysname", "release", "machine")], "\n")
cat("sf capabilities:\n")
print(sf::sf_extSoftVersion())

# Test lubridate time parsing (used internally by CopernicusMarine)
cat("\nlubridate parsing tests:\n")
cat("ymd '1998-01-01':", format(lubridate::ymd("1998-01-01")), class(lubridate::ymd("1998-01-01")), "\n")
cat("ymd '2018-12-31':", format(lubridate::ymd("2018-12-31")), class(lubridate::ymd("2018-12-31")), "\n")
cat("as.numeric(ymd('1998-01-01')):", as.numeric(lubridate::ymd("1998-01-01")), "\n")
cat("as.numeric(ymd('2018-12-31')):", as.numeric(lubridate::ymd("2018-12-31")), "\n")
cat("as.numeric(ymd_hms('1998-01-01T00:00:00Z')):", as.numeric(lubridate::ymd_hms("1998-01-01T00:00:00Z")), "\n")
cat("as.numeric(ymd_hms('2018-12-31T00:00:00Z')):", as.numeric(lubridate::ymd_hms("2018-12-31T00:00:00Z")), "\n")
```

```{r}
study_bbox <- sf::st_bbox(
  c(xmin = 3, ymin = 42.5, xmax = 7, ymax = 44),
  crs = 4326
)
```

### Test 1: Yearly layer with timerange (failing on CI)

```{r}
#| error: true
result1 <- cms_download_subset(
  product = "MEDSEA_MULTIYEAR_PHY_006_004",
  layer = "cmems_mod_med_phy-tem_my_4.2km_P1Y-m",
  variable = "bottomT",
  region = study_bbox,
  timerange = c("1998-01-01", "2018-12-31")
)
result1
cat("Test 1 dims:", dim(result1), "\n")
```

### Test 2: Yearly layer with time steps matching Jan 1

```{r}
#| error: true
result2 <- cms_download_subset(
  product = "MEDSEA_MULTIYEAR_PHY_006_004",
  layer = "cmems_mod_med_phy-tem_my_4.2km_P1Y-m",
  variable = "bottomT",
  region = study_bbox,
  timerange = c("1998-01-01", "2018-01-01")
)
result2
cat("Test 2 dims:", dim(result2), "\n")
```

### Test 3: Yearly layer without timerange

```{r}
#| error: true
result3 <- cms_download_subset(
  product = "MEDSEA_MULTIYEAR_PHY_006_004",
  layer = "cmems_mod_med_phy-tem_my_4.2km_P1Y-m",
  variable = "bottomT",
  region = study_bbox
)
result3
cat("Test 3 dims:", dim(result3), "\n")
```

### Test 5: No timerange + filter to sampled years via as.POSIXct

```{r}
#| error: true
time_vals <- stars::st_get_dimension_values(result3, "time")
cat("time_vals class:", class(time_vals), "\n")
cat("First value:", time_vals[1], "\n")

# Convert to POSIXct to handle both POSIXct and udunits refsys
time_vals <- as.POSIXct(time_vals)
cat("After as.POSIXct class:", class(time_vals), "\n")
cat("First value:", format(time_vals[1]), "\n")

sampled_years <- c(1998, 2004, 2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2015, 2018)
keep <- as.integer(format(time_vals, "%Y")) %in% sampled_years
cat("Keep:", sum(keep), "of", length(keep), "\n")
cat("Years kept:", sort(unique(as.integer(format(time_vals[keep], "%Y")))), "\n")

filtered <- result3[,,, keep]
cat("Filtered dims:", dim(filtered), "\n")
```

### Test 4: Monthly layer with timerange

```{r}
#| error: true
result4 <- cms_download_subset(
  product = "MEDSEA_MULTIYEAR_PHY_006_004",
  layer = "cmems_mod_med_phy-temp_my_4.2km_P1M-m",
  variable = "bottomT",
  region = study_bbox,
  timerange = c("1998-01-01", "2018-12-31")
)
result4
cat("Test 4 dims:", dim(result4), "\n")
```
