---
title: "4: Characterising Benthic Biozones"
subtitle: "Combining Habitat Maps, Bathymetry, Temperature, Species Occurrences and Body Size Data in the Gulf of Lion"
difficulty: "Advanced"
time: "~90 minutes"
data-type: "WFS + WCS + External sources"
description: "Learn to integrate EMODnet WFS and WCS services with external data sources including the Copernicus Marine Service and trait databases. Characterise benthic biozones in the Gulf of Lion using habitat composition, substrate type, environmental conditions, and body size distributions."
datasets: "Seabed habitat polygons, benthic species occurrences, bathymetry, bottom temperature, body size data"
df-print: paged
bibliography: references.bib
format:
  html:
    embed-resources: false
    code-link: true
---

## Introduction

The Gulf of Lion, located in the northwestern Mediterranean off the coast of southern France, hosts a rich diversity of benthic communities shaped by the Rhône River delta, diverse substrates from sandy beaches to deep canyons, and strong environmental gradients. Characterising the biozones of this region, from the sunlit infralittoral to the deep bathyal, requires bringing together habitat maps, environmental data, species occurrences, and functional traits such as body size.

In this tutorial, we demonstrate how to **integrate multiple EU marine data infrastructures** to characterise benthic biozones in the Gulf of Lion. We combine:

-   **EMODnet WFS** for habitat maps and species occurrence records
-   **EMODnet WCS** for bathymetry data
-   **Copernicus Marine Service** for bottom temperature
-   **MOBS database** for species body size data

This multi-source approach reflects real-world marine ecological analysis, where insights emerge from combining data across platforms and domains.

### Why Include Body Size?

Body size is a fundamental trait linked to metabolism, feeding strategy, life history, and ecological role. Adding body size to a biozone characterisation enriches the picture beyond habitat composition and physical conditions. The distribution of body sizes across biozones can reveal:

-   **Environmental filtering**: Temperature and depth gradients may favour certain body size ranges
-   **Habitat associations**: Different seabed types may support communities with distinct size structures
-   **Community function**: Body size distributions affect energy flow and ecosystem processes

### Learning Objectives

By the end of this tutorial, you will be able to:

-   Retrieve seabed habitat polygons from EMODnet Seabed Habitats WFS
-   Access benthic species occurrence data from EMODnet Biology WFS
-   Download bathymetry rasters from EMODnet Bathymetry WCS
-   Access bottom temperature data from the Copernicus Marine Service
-   Link species records to body size data from external databases
-   Use spatial joins to assign habitat and biozone attributes to species occurrences based on location
-   Join datasets using shared identifiers (AphiaIDs)
-   Extract zonal statistics from rasters using habitat polygons
-   Characterise benthic biozones using habitat, environmental, and trait data
-   Create interactive maps with the `leaflet` package

### Data Sources

**EMODnet WFS (Vector Data):**

| Layer | Service | Description |
|------------------|----------------------|--------------------------------|
| `eusm2025_eunis2019_full` | seabed_habitats_general_datasets_and_products | EU Seabed Map with EUNIS 2019 classification |
| `eb185_benthos_full_matched` | biology | Benthic species occurrence records from European seas |

**EMODnet WCS (Raster Data):**

| Coverage             | Service    | Description                   |
|----------------------|------------|-------------------------------|
| `emodnet__mean_2022` | bathymetry | Mean water depth (bathymetry) |

**External Sources:**

| Source | Description |
|----------------------------|--------------------------------------------|
| [Copernicus Marine Service](https://marine.copernicus.eu/) | Bottom temperature from [MEDSEA_MULTIYEAR_PHY_006_004](https://data.marine.copernicus.eu/product/MEDSEA_MULTIYEAR_PHY_006_004/description) reanalysis |
| [MOBS Database](https://github.com/crmcclain/MOBS_OPEN) | Marine Organism Body Size database [@mccallum2025] |

## Setup

### Packages

This tutorial uses a broader set of packages than previous tutorials, reflecting the multi-source integration workflow:

```{r}
#| label: load-packages
#| message: false
library(emodnet.wfs)
library(emodnet.wcs)
library(CopernicusMarine)
library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(ggridges)
library(patchwork)
library(tidyterra)
library(leaflet)
library(htmlwidgets)

```

```{r}
#| label: load-internal-packages
#| include: false
library(emodnet.bio.r.geo.tutorials)
```

::: {.callout-tip collapse="true"}
## Package installation

Most packages are available from CRAN:

``` r
install.packages(c(
  "emodnet.wfs", "CopernicusMarine", "terra", "sf", "stars",
  "dplyr", "tidyr", "purrr", "readr", "stringr", "ggplot2",
  "ggridges", "janitor", "patchwork", "tidyterra", "leaflet",
  "htmlwidgets",
  "rnaturalearth"
))
```

The `emodnet.wcs` package is not yet on CRAN; install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("EMODnet/emodnet.wcs")
```
:::

### Study Area

We focus on the Gulf of Lion in the northwestern Mediterranean. This region offers good coverage across most data sources involved, with diverse substrate types (sand, mud, coarse sediments, seagrass meadows) and spatial sampling across 165 unique locations. The Gulf of Lion also spans multiple biozones from infralittoral to bathyal depths.

We start by defining a bounding box for our study area in WGS 84 (longitude/latitude degrees), which we will use to filter data downloads to this region. We also create an EPSG:3035 version, an equal-area projection that several of our data sources use (we discuss this CRS in more detail in the [spatial data integration](#spatial-data-integration) section):

```{r}
#| label: define-study-area-bbox
# Study area bounding box in WGS 84 (degrees)
study_bbox <- sf::st_bbox(
  c(xmin = 3, ymin = 42.5, xmax = 7, ymax = 44),
  crs = 4326
)
# As polygon geometries for visualisation and cropping
bbox_polygon <- sf::st_as_sfc(study_bbox)

# EPSG:3035 bbox and polygon
bbox_polygon_3035 <- sf::st_transform(bbox_polygon, 3035)
study_bbox_3035 <- sf::st_bbox(bbox_polygon_3035)
```

```{r}
#| label: fig-study-area
#| code-fold: true
#| fig-cap: "Study area: Gulf of Lion in the northwestern Mediterranean"

# Get Mediterranean coastlines
med_countries <- rnaturalearth::ne_countries(
  scale = "medium",
  returnclass = "sf"
) |>
  sf::st_crop(c(xmin = -2, ymin = 38, xmax = 12, ymax = 46)) |>
  sf::st_transform(3035)

ggplot() +
  geom_sf(
    data = med_countries,
    fill = "gray90",
    color = "gray60",
    linewidth = 0.3
  ) +
  geom_sf(data = bbox_polygon, fill = NA, color = "#004494", linewidth = 1.5) +
  coord_sf(crs = 4326, xlim = c(-2, 12), ylim = c(38, 46)) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(panel.grid = element_line(color = "gray85", linewidth = 0.2))
```

## Retrieving Seabed Habitat Data

Our first data source is seabed habitat polygons from EMODnet Seabed Habitats. These polygons define the spatial framework for our biozone characterisation, providing EUNIS habitat codes, substrate types, and biozone classifications across the study area.

### Connecting to EMODnet Seabed Habitats WFS

For an introduction to working with WFS services and the `emodnet.wfs` package, see [Tutorial 1](tutorial-01.qmd). Here we apply the same workflow to the Seabed Habitats service.

```{r}
#| label: init-habitats-wfs
habitat_wfs <- emodnet_init_wfs_client(
  service = "seabed_habitats_general_datasets_and_products"
)
```

### Exploring Habitat Layers

We need habitat polygons from the [EUSeaMap](https://emodnet.ec.europa.eu/en/seabed-habitats), the broad-scale predictive habitat map for European seas. The service contains many layers, so we filter by title to find the EUSeaMap products:

```{r}
#| label: explore-habitat-layers
#| cache: true
habitat_layers <- emodnet_get_wfs_info(habitat_wfs)

# Search for EUSeaMap layers by title
habitat_layers |>
  filter(grepl("EUSeaMap", title)) |>
  select(layer_name, title)
```

The results show separate layers for energy, substrate, biozone, and two habitat classification systems (EUNIS 2007 and EUNIS 2019). We choose `eusm2025_eunis2019_full` because EUNIS 2019 is the current classification standard and this layer includes both the habitat code and the substrate and biozone attributes we need.

### Understanding Habitat Classifications

Let's examine the attribute structure of the habitat layer:

```{r}
#| label: inspect-habitat-attributes
#| cache: true
habitat_attrs <- layer_attribute_descriptions(
  wfs = habitat_wfs,
  layer = "eusm2025_eunis2019_full"
)

habitat_attrs
```

Key attributes include:

-   **`geom`**: Geometry column. WFS layers can use different names for this column, so we extract it programmatically below to construct CQL bounding box filters
-   **`eunis2019c`**: EUNIS 2019 habitat classification (level 3)
-   **`eunis2019d`**: Human-readable EUNIS 2019 habitat description
-   **`substrate`**: Seabed substrate type
-   **`biozone`**: Biological zone

Let's check what habitat values are present across the full layer (not just our study area) using `layer_attribute_inspect()`:

```{r}
#| label: explore-habitat-types
#| cache: true
eunis_values <- layer_attribute_inspect(
  wfs = habitat_wfs,
  layer = "eusm2025_eunis2019_full",
  attribute = "eunis2019c"
)

eunis_values
```

Scrolling through the values, you will notice that not all entries are clean EUNIS codes. Some contain `"Na"` (polygons with no habitat classification) and others contain `"or"` (e.g. `"MB551 or MB552"`), representing polygons where the classification is ambiguous between two habitat types. Let's quantify how many polygons across the entire layer are affected:

```{r}
#| label: count-invalid-habitat-values
eunis_values |>
  filter(.data[["."]] == "Na" | grepl("or", .data[["."]]))
```

These are data gaps in the EUSeaMap product itself, not specific to our study area. Since these values are not valid EUNIS habitat classifications, we filter them out server-side using CQL to keep our habitat dataset clean.

### Building an Attribute CQL Filter

In previous tutorials we used CQL (Common Query Language) filters to spatially subset WFS data by bounding box. CQL can also filter on **attribute values**, which we use here for the first time to exclude the problematic `eunis2019c` entries identified above. We combine both spatial and attribute conditions into a single filter so the server only returns the data we need.

We start by extracting the geometry column name for the [BBOX filter](tutorial-01.qmd#filtering-by-location-pre-download):

```{r}
#| label: habitat-geom-col
habitat_geom_col <- habitat_attrs$name[habitat_attrs$geometry]
```

We also want to combine the spatial BBOX filter with **attribute filters** to exclude the problematic `eunis2019c` values we identified above. CQL supports combining conditions with `AND`: we use `<>` to exclude exact matches and `NOT LIKE` with `%` wildcards for pattern matching. For more on attribute filtering syntax, see the [emodnet.wfs ECQL filtering vignette](https://emodnet.github.io/emodnet.wfs/articles/ecql_filtering.html#filtering-by-attributes).

```{r}
#| label: habitat-cql-filter
# Spatial filter: bounding box
bbox_filter <- sprintf(
  "BBOX(%s, %s, %s, %s, %s, 'EPSG:4326')",
  habitat_geom_col,
  study_bbox["xmin"],
  study_bbox["ymin"],
  study_bbox["xmax"],
  study_bbox["ymax"]
)

# Attribute filters: exclude invalid EUNIS codes
na_filter <- "eunis2019c <> 'Na'"
or_filter <- "eunis2019c NOT LIKE '%or%'"

# Combine all conditions
habitat_cql_filter <- paste(bbox_filter, na_filter, or_filter, sep = " AND ")
habitat_cql_filter
```

The `<>` operator excludes polygons where `eunis2019c` is exactly `"Na"`, and `NOT LIKE '%or%'` excludes any value containing the substring `"or"` (the `%` characters are CQL wildcards matching any sequence of characters).

### Downloading Habitat Polygons

Now we download with the combined spatial and attribute filter. We use [GeoJSON format](tutorial-01.qmd#why-use-geojson-format) (`outputFormat = "application/json"`) for faster downloads and simpler geometry types, and set `simplify = TRUE` to return a single `sf` data frame rather than a list of layers. The habitat layer's native CRS is Web Mercator (EPSG:3857), but we request **EPSG:3035** (ETRS89-extended / LAEA Europe) via `crs = 3035`. This is an [equal-area projection](../concepts.qmd#why-this-matters-for-marine-spatial-analysis) well suited to European marine data: every cell and polygon covers the same ground area regardless of latitude, which matters for zonal statistics and spatial joins later in the tutorial:

```{r}
#| label: download-habitat-polygons
#| cache: true
seabed_habitats <- emodnet_get_layers(
  wfs = habitat_wfs,
  layers = "eusm2025_eunis2019_full",
  cql_filter = habitat_cql_filter,
  outputFormat = "application/json",
  simplify = TRUE,
  crs = 3035
)

nrow(seabed_habitats)
```

### Cropping Polygons to the Study Area

The WFS server returns any polygon that overlaps our bounding box, but it returns each polygon in its entirety. Some polygons, particularly large deep-sea habitat types like the abyssal plain, may extend far beyond our study area. We can see this by checking the bounding box of the downloaded data:

```{r}
#| label: check-habitat-bbox
# Transform to WGS 84 to show extent in degrees
st_bbox(seabed_habitats) |> st_transform(4326)
```

The extent (`r round(st_bbox(seabed_habitats) |> st_transform(4326))[c("xmin","xmax")] |> paste(collapse="--")`°E, `r round(st_bbox(seabed_habitats) |> st_transform(4326))[c("ymin","ymax")] |> paste(collapse="--")`°N) is much larger than our study area (`r study_bbox["xmin"]`--`r study_bbox["xmax"]`°E, `r study_bbox["ymin"]`--`r study_bbox["ymax"]`°N). Leaving these oversized polygons in place causes two problems: it can skew analyses by including data outside the area of interest, and later, when we extract zonal statistics from rasters constrained to our bounding box, the overhanging parts of polygons produce `NaN` values (the raster has no cells to average over those areas).

We fix this with `sf::st_crop()`, which clips all geometries to a target rectangle. We use `bbox_polygon_3035`, the EPSG:3035 version of our study area polygon created at the start of the tutorial:

```{r}
#| label: crop-habitat-polygons
seabed_habitats <- seabed_habitats |>
  st_crop(bbox_polygon_3035)

st_bbox(seabed_habitats) |> st_transform(4326)
```

The bounding box of our seabed habitat data now matches our study area.

### Inspecting Downloaded Data

Let's preview the key columns. Each row is a polygon with a EUNIS habitat code, a human-readable description, substrate type, and biozone:

```{r}
#| label: preview-habitat-polygons
seabed_habitats |>
  select(eunis2019c, eunis2019d, substrate, biozone) |>
  head(10)
```

::: {.callout-note}
The WFS layer may use a different name for its geometry column (e.g. `the_geom`, `msGeometry`), but after download the `sf` package standardises the column name to `geometry`. This is normal and applies to all WFS downloads throughout this tutorial.
:::

The geometry column contains sf `r unique(st_geometry_type(seabed_habitats))` features. Each habitat feature is a polygon (or collection of polygons) delineating an area of seabed with a particular classification. The columns show:

- **`eunis2019c`**: the EUNIS 2019 level 3 code (e.g. `MB552`)
- **`eunis2019d`**: the human-readable habitat description
- **`substrate`**: the physical substrate (e.g. Sand, Mud, Coarse substrate)
- **`biozone`**: the depth zone (e.g. Infralittoral, Deep circalittoral)

The `biozone` variable represents depth zones from shallow (Infralittoral) to deep (Abyssal). 

```{r}
#| label: show-biozone
unique(seabed_habitats$biozone)
```

We convert it to an ordered factor for meaningful analysis along depth gradients, and standardise `substrate` casing (the source data mixes "SAND" and "Sand"):

```{r}
#| label: order-biozone
biozone_levels <- c(
  "Infralittoral",
  "Shallow circalittoral",
  "Deep circalittoral",
  "Upper bathyal",
  "Lower bathyal",
  "Abyssal"
)

seabed_habitats <- seabed_habitats |>
  mutate(
    biozone = factor(biozone, levels = biozone_levels, ordered = TRUE),
    substrate = str_to_sentence(substrate)
  )
```

We can confirm the CQL filter worked by checking the unique `eunis2019c` values in the downloaded data:

```{r}
#| label: verify-clean-eunis
sort(unique(seabed_habitats$eunis2019c))
```

All clean EUNIS codes, with no `"Na"` or `"or"` values. Let's see which habitat types are most common in our study area:

```{r}
#| label: habitat-type-summary
# Summarise habitat types present
seabed_habitats |>
  st_drop_geometry() |>
  count(eunis2019c, sort = TRUE) |>
  head(10)
```

### Visualising Seabed Habitats

```{r}
#| label: fig-seabed-habitats
#| code-fold: true
#| fig-cap: "Seabed habitat classification in the Gulf of Lion (EUNIS 2019)"
#| fig-height: 10
#| fig-width: 8

ggplot() +
  geom_sf(data = med_countries, fill = "gray90", colour = "gray70") +
  geom_sf(
    data = seabed_habitats,
    aes(fill = str_wrap(eunis2019d, width = 50)),
    colour = NA,
    alpha = 0.8
  ) +
  coord_sf(
    xlim = c(
      study_bbox_3035["xmin"],
      study_bbox_3035["xmax"]
    ),
    ylim = c(
      study_bbox_3035["ymin"],
      study_bbox_3035["ymax"]
    )
  ) +
  scale_fill_viridis_d(option = "turbo", name = NULL) +
  labs(title = "Seabed Habitats (EUNIS 2019)") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.key.size = unit(0.4, "cm"),
    plot.margin = margin(2, 2, 2, 2)
  ) +
  guides(fill = guide_legend(ncol = 2, override.aes = list(alpha = 1)))
```

## Retrieving Species Occurrence Data

Next, we retrieve benthic species occurrence records from EMODnet Biology. These are geolocated records whose species names have also been matched to the World Register of Marine Species (WoRMS) via an AphiaID code, ensuring taxonomic consistency. Later, we will use a spatial join to link these point occurrences to the habitat polygons we just downloaded, assigning each record to a habitat type, biozone and substrate based on where it falls on the seabed map.

::: {.callout-tip collapse="true"}
## Taxonomic matching with the `worrms` package

The data we use here are already matched to WoRMS, but if you need to match your own species lists to the WoRMS taxonomic backbone, the [`worrms`](https://cran.r-project.org/package=worrms) R package provides a programmatic interface to WoRMS. It supports fuzzy name matching, retrieval of accepted names and synonyms, and lookup of taxonomic hierarchies, useful for standardising species names across datasets.
:::

### Connecting to EMODnet Biology WFS

We follow the same discovery workflow as for the habitat service above.

```{r}
#| label: init-biology-wfs
bio_wfs <- emodnet_init_wfs_client(service = "biology")
```

### Exploring Available Layers

Let's search for benthic occurrence data:

```{r}
#| label: explore-biology-layers
#| cache: true
bio_layers <- emodnet_get_wfs_info(bio_wfs)

# Filter for benthos-related layers
bio_layers |>
  filter(grepl("benthos", layer_name, ignore.case = TRUE)) |>
  select(layer_name, title, abstract) |>
  # Keep first paragraph of abstract for display
  mutate(abstract = str_extract(abstract, "^[^\n]+")) |>
  knitr::kable()
```

We choose `eb185_benthos_full_matched` because it contains individual geolocated occurrence records, each matched to WoRMS (providing AphiaIDs we can use to link to trait data). The `benthos_europe` layer, by contrast, provides gridded presence/absence summaries rather than individual occurrence records, so it is less suited to point-level analysis.

### Understanding Layer Structure

Before downloading, let's examine what attributes are available (see [Tutorial 1](tutorial-01.qmd#understanding-layer-structure) for more on inspecting WFS layer metadata):

```{r}
#| label: inspect-benthos-attributes
#| cache: true
benthos_attrs <- layer_attribute_descriptions(
  wfs = bio_wfs,
  layer = "eb185_benthos_full_matched"
)

benthos_attrs
```

Key attributes for our analysis include:

-   **`st_point`**: Geometry column (extracted programmatically for the CQL filter, as with the habitat layer above)
-   **`aphiaid`**: WoRMS identifier for linking to trait databases
-   **`scientificname`**: Species name for display
-   **`lon`**, **`lat`**: Coordinates
-   **`year`**: Sampling year
-   **`eusm_polygon_id`**: Link to the EUSeaMap grid polygon
-   **`EUNIS2019C`**, **`EUNIS2019D`**: EUNIS 2019 habitat classification (level 3 and 4), pre-matched via the EUSeaMap grid. We will not use these for joining; instead we use a spatial join to assign habitat attributes based on where each occurrence falls within our habitat polygons

### Downloading Occurrence Data

As with the habitat data, we construct a CQL BBOX filter for our study area:

```{r}
#| label: define-benthos-bbox-filter
benthos_geom_col <- benthos_attrs$name[benthos_attrs$geometry]

benthos_bbox_filter <- sprintf(
  "BBOX(%s, %s, %s, %s, %s, 'EPSG:4326')",
  benthos_geom_col,
  study_bbox["xmin"],
  study_bbox["ymin"],
  study_bbox["xmax"],
  study_bbox["ymax"]
)
```

Now we download the data with `emodnet_get_layers()`, passing our CQL filter to restrict the query to the study area. As with the habitat data, we use GeoJSON format and request EPSG:3035 so both vector datasets share the same equal-area CRS:

```{r}
#| label: download-benthos-occurrences
#| cache: true
benthos_occurrences <- emodnet_get_layers(
  wfs = bio_wfs,
  layers = "eb185_benthos_full_matched",
  cql_filter = benthos_bbox_filter,
  outputFormat = "application/json",
  simplify = TRUE,
  crs = 3035
)

# Check the result
nrow(benthos_occurrences)
```

### Inspecting Downloaded Data

Each record is a geolocated `POINT` (in contrast to the habitat **polygons** we retrieved earlier), representing a species observed at a specific location and time. The key columns for our analysis are:

-   **`aphiaid`**: WoRMS identifier, which we will use to link to body size data
-   **`scientificname`**: Species name
-   **`year`**: Sampling year
-   **`geometry`**: Location of the observation

```{r}
#| label: preview-benthos
benthos_occurrences |>
  select(aphiaid, scientificname, year) |>
  head(10)
```

Let's examine the taxonomic diversity and data completeness:

```{r}
#| label: benthos-summary
# How many unique species?
n_species <- n_distinct(benthos_occurrences$aphiaid)

# How many records have valid AphiaIDs for trait matching?
n_with_aphiaid <- sum(!is.na(benthos_occurrences$aphiaid))

cat(sprintf(
  "Total records: %d\nUnique species (AphiaIDs): %d\nRecords with valid AphiaID: %d (%.1f%%)\n",
  nrow(benthos_occurrences),
  n_species,
  n_with_aphiaid,
  100 * n_with_aphiaid / nrow(benthos_occurrences)
))
```

### Quick Visualisation

```{r}
#| label: fig-benthos-occurrences
#| code-fold: true
#| fig-cap: "Benthic species occurrence records in the Gulf of Lion"

ggplot() +
  geom_sf(data = med_countries, fill = "gray90", colour = "gray70") +
  geom_sf(
    data = benthos_occurrences,
    colour = "#0077BE",
    size = 2,
    alpha = 0.3
  ) +
  coord_sf(
    xlim = c(
      study_bbox_3035["xmin"],
      study_bbox_3035["xmax"]
    ),
    ylim = c(
      study_bbox_3035["ymin"],
      study_bbox_3035["ymax"]
    )
  ) +
  labs(
    title = "Benthic Species Occurrences",
    subtitle = sprintf(
      "%d records in the Gulf of Lion",
      nrow(benthos_occurrences)
    )
  ) +
  theme_minimal()
```

## Retrieving Bathymetry Data

Bathymetry (water depth) provides essential environmental context for our habitat characterisation. Unlike the vector data (polygons and points) we have downloaded so far, bathymetry is served as **raster data**: a grid of cells where each cell holds a depth value. We retrieve this from EMODnet Bathymetry using [WCS](../concepts.qmd#wcs-web-coverage-service) (see [Tutorial 2](tutorial-02.qmd) for an introduction to working with the `emodnet.wcs` package).

### Connecting to EMODnet Bathymetry WCS

```{r}
#| label: init-bathymetry-wcs
bathy_wcs <- emdn_init_wcs_client(service = "bathymetry")
```

### Exploring Available Coverages

`emdn_get_coverage_ids()` lists all available coverage layers from the service:

```{r}
#| label: list-bathymetry-coverages
bathy_coverage_ids <- emdn_get_coverage_ids(bathy_wcs)
bathy_coverage_ids
```

The coverage IDs encode the product type and reference year. The `mean` coverages provide the gridded mean depth, with year suffixes indicating different releases of the bathymetric compilation. We use the most recent release (`emodnet__mean_2022`). `emdn_get_coverage_info()` provides metadata including the spatial resolution, extent, and coordinate reference system:

```{r}
#| label: get-bathymetry-info
bathy_info <- emdn_get_coverage_info(
  wcs = bathy_wcs,
  coverage_ids = "emodnet__mean_2022"
)

bathy_info |>
  knitr::kable()
```

The resolution is approximately 1/960 of a degree (~115 m at this latitude). Note that the coverage CRS is EPSG:4326, which is why we pass `study_bbox` (our WGS 84 bounding box) when downloading. The nil value is `NA`, so no special handling of missing data codes is needed.

### Downloading Bathymetry

We download the bathymetry raster for our study area using `emdn_get_coverage()`, which returns a `SpatRaster`:

```{r}
#| label: download-bathymetry
bathymetry <- emdn_get_coverage(
  wcs = bathy_wcs,
  coverage_id = "emodnet__mean_2022",
  bbox = study_bbox
)

bathymetry
```

### Inspecting Downloaded Data

EMODnet bathymetry uses the convention of **negative values for depth below sea level** and positive values for elevation above it. Since our bounding box overlaps the coastline, the raster includes both marine cells (negative) and land cells (positive). We can check the value range with `terra::global()`, which computes a summary statistic across all cells of a `SpatRaster`:

```{r}
#| label: check-bathymetry-range
terra::global(bathymetry, range, na.rm = TRUE)
```

The negative minimum confirms deep marine areas. The positive maximum (~1,900 m) may seem surprising for a marine study area, but our bounding box extends to 44°N, which reaches well inland over southern France into the foothills of the Pyrenees and Massif Central. EMODnet bathymetry is a digital terrain model that covers both sea and land, so these terrestrial elevations are included in the download. We set land and sea-level cells (values >= 0) to `NA`, then negate the remaining values so that depth is expressed as a positive number (deeper = larger):

```{r}
#| label: process-bathymetry
bathymetry[bathymetry >= 0] <- NA
depth <- -1 * bathymetry
names(depth) <- "depth_m"
```

### Visualising Bathymetry

```{r}
#| label: fig-bathymetry
#| code-fold: true
#| fig-cap: "Bathymetry of the Gulf of Lion (depth in metres)"

ggplot() +
  geom_spatraster(data = depth) +
  scale_fill_viridis_c(
    name = "Depth (m)",
    na.value = "gray30",
    option = "mako",
    direction = -1,
    breaks = c(0, 200, 500, 1000, 2000, 2500),
    # Allocate more of the colour gradient to shallow depths so that
    # shelf features and canyon heads are visible despite the large
    # overall depth range (~0–2700 m)
    values = scales::rescale(
      c(0, 50, 200, 500, 1000, 2000, 2500),
      to = c(0, 1),
      from = c(0, 2500)
    )
  ) +
  labs(title = "Gulf of Lion Bathymetry") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(
    fill = guide_colorbar(
      barwidth = 15,
      barheight = 0.5,
      title.position = "top"
    )
  )
```

## Retrieving Bottom Temperature

Bottom temperature provides insight into thermal niches and is a key variable for characterising environmental conditions across biozones. We access this from the [Copernicus Marine Service](https://marine.copernicus.eu/) (CMEMS), the marine component of the EU's [Copernicus](https://www.copernicus.eu/) Earth observation programme. Copernicus is the Earth observation component of the EU's Space programme, drawing on satellite and in-situ observations to monitor the planet. All Copernicus data and products are provided on a free, full and open basis under EU regulation. CMEMS provides ocean analysis, forecast and reanalysis products covering the global ocean and European regional seas. We interact with it in R using the [`CopernicusMarine`](https://pepijn-devries.github.io/CopernicusMarine/) package.

### How Copernicus Marine Service data is organised

The EMODnet services we used above follow [OGC standards](../concepts.qmd#ogc-standards-and-spatial-data-types): WFS serves vector features and WCS serves raster coverages, each accessed through a service client. Copernicus Marine Service organises its data differently. Data is stored in both NetCDF and [zarr](../concepts.qmd#zarr-.zarr) (a cloud-native array format) and discovered through a [STAC](https://stacspec.org/) (SpatioTemporal Asset Catalog) catalogue. The `CopernicusMarine` package handles this transparently (using the zarr stores for efficient spatial subsetting), but the terminology differs from the OGC services:

| EMODnet | Copernicus Marine | Copernicus Example |
|---------|-------------------|--------------------|
| Service (e.g., `"biology"`) | **Product** (a dataset with a unique ID) | `MEDSEA_MULTIYEAR_PHY_006_004` |
| Layer (e.g., `"eb185_benthos_full_matched"`) | **Layer** (a specific temporal resolution or subset) | `cmems_mod_med_phy-temp_my_4.2km_P1M-m` |
| Attributes (WFS) or Bands (WCS) | **Variables** (physical quantities stored in the array) | `bottomT`, `thetao` |
| 2D features or grids | **Multi-dimensional arrays** with longitude, latitude, depth, and time | 4D datacube |

Because the data is multi-dimensional, `cms_download_subset()` returns a [`stars`](../concepts.qmd#stars---spatiotemporal-arrays) object rather than the `sf` or `terra` objects returned by EMODnet services. For more background, see the [stars](../concepts.qmd#stars---spatiotemporal-arrays) and [zarr](../concepts.qmd#zarr-.zarr) sections in the Geospatial Concepts page.

The metadata returned by `CopernicusMarine` functions follows the STAC specification. For a general introduction to what STAC is and how catalogues, collections, and items relate, see the [STAC intro tutorial](https://stacspec.org/en/tutorials/intro-to-stac/). In practice, the deeply nested metadata structures can be challenging to navigate, but contain rich information about layers, variables, dimensions, and time ranges, as we explore below.

The discovery workflow follows a similar pattern to EMODnet: list what is available, inspect metadata, then download a subset.

### Choosing a Data Product

Bottom temperature is typically available in physical reanalysis products, identified by `MULTIYEAR_PHY` in their product IDs. `cms_products_list()` lists all available products; we filter to physical reanalysis:

```{r}
#| label: explore-copernicus-products
#| cache: true
# List physical reanalysis products
cms_products_list() |>
  filter(grepl("MULTIYEAR_PHY", product_id)) |>
  select(product_id, title) |>
  knitr::kable()
```

A Copernicus **product** plays the same role as an EMODnet **service** (like `"biology"` or `"bathymetry"`): it is a thematic collection that groups related layers together. For small study areas, regional products typically offer higher resolution and are specifically validated for that region. For our Gulf of Lion study, the Mediterranean product ([MEDSEA_MULTIYEAR_PHY_006_004](https://data.marine.copernicus.eu/product/MEDSEA_MULTIYEAR_PHY_006_004/description)) provides \~4.2 km resolution compared to \~8 km for the global product.

### Browsing Available Variables

Unlike OGC services where each layer contains a single dataset, Copernicus products are built on a datacube architecture. A single product can bundle multiple physical variables (temperature, salinity, currents) across multiple layers. We can get a high-level overview of all variables available across the entire product using `cms_product_details()`:

```{r}
#| label: explore-med-product-details
#| cache: true
med_product <- cms_product_details("MEDSEA_MULTIYEAR_PHY_006_004")

# allVariables aggregates human-readable variable names across ALL layers
med_product$properties$allVariables |> unlist()
```

This confirms the product includes "Sea water potential temperature at sea floor", which is what we need. However, this is a product-wide summary: it aggregates variable names across all layers, so not every variable listed here is necessarily available in every layer. To find out exactly which variables are available in which layers, and to get the variable `id`s needed for downloading, we need to drill into the layer-level metadata.

### Discovering Layers and Their Variables

A product contains multiple **layers**, each offering a different temporal resolution of the same underlying data (e.g. daily vs monthly means). `cms_product_services()` lists them:

```{r}
#| label: explore-med-services
#| cache: true
med_services <- cms_product_services("MEDSEA_MULTIYEAR_PHY_006_004")

# Each row is a layer
med_services$id
```

The layer `id`s encode the variable group and temporal resolution (e.g., `temp` for temperature, `P1M-m` for monthly means, `P1D-m` for daily). Version suffixes (e.g., `_202511`) can be omitted when downloading.

Each layer carries rich variable metadata in its `cube:variables` property. The [STAC datacube extension](https://github.com/stac-extensions/datacube) defines a base set of fields (`dimensions`, `type`, `unit`, `description`), and Copernicus Marine adds its own fields on top. Let's examine the structure by drilling into the first variable of the first layer. `med_services` is a data frame with list-columns, where each row is a layer. `properties[[1]]` accesses the STAC metadata for the first layer, `cube:variables` is a named list of that layer's variables, and `[[1]]` selects the first variable entry:

```{r}
#| label: inspect-cube-variables
# Examine the metadata structure of the first variable in the first layer
str(med_services$properties[[1]]$`cube:variables`[[1]], max.level = 2)
```

The standard STAC datacube fields include `dimensions` (which axes the variable spans), `type` (`"data"` for measured values), and `unit`. In standard STAC, each variable is identified by its key name in the `cube:variables` map. Copernicus Marine also stores this as an explicit `id` field inside each variable object, along with other Copernicus-specific additions: `name$en` (a human-readable description), `standardName` (the [CF convention](https://cfconventions.org/Data/cf-standard-names/current/build/cf-standard-name-table.html) name), and visualisation hints like `colormapId` and value ranges.

We can use these fields to build a clear mapping of layers to their variables. Since we need bottom temperature, we filter to the temperature layers:

```{r}
#| label: check-temp-layer-variables
# Extract variable details from each temperature layer
med_services |>
  filter(grepl("tem", id, ignore.case = TRUE)) |>
  rowwise() |>
  reframe(
    layer = id,
    variable = purrr::map_chr(properties$`cube:variables`, "id"),
    description = purrr::map_chr(properties$`cube:variables`, \(x) x$name$en),
    unit = purrr::map_chr(properties$`cube:variables`, "unit")
  ) |>
  knitr::kable()
```

All temperature layers contain `bottomT` (sea floor potential temperature) and `thetao` (potential temperature through the water column). For comprehensive documentation of all variable names, layer naming conventions, and production methodology, see the [Product User Manual (PUM)](https://documentation.marine.copernicus.eu/PUM/CMEMS-MED-PUM-006-004.pdf).

We choose the yearly means layer (`P1Y-m`) rather than daily (`P1D-m`) or monthly (`P1M-m`). The pre-computed climatology might seem ideal, but it provides a long-term average across all years, whereas we want to subset to the specific years when our species were sampled. Yearly means let us do exactly that while keeping the download small.

```{r}
#| label: select-temp-layer
# Select the yearly temperature layer
temp_yearly <- med_services |>
  filter(id == "cmems_mod_med_phy-tem_my_4.2km_P1Y-m_202211")
```

The `cube:dimensions` property tells us the available **temporal extent** and **time steps** for our chosen layer, which is essential for knowing what date range we can request:

```{r}
#| label: check-temp-layer-time
time_dim <- temp_yearly$properties[[1]]$`cube:dimensions`$time

# Temporal extent (start and end dates)
cat("Temporal extent:", time_dim$extent[[1]], "to", time_dim$extent[[2]])
cat("Total time steps:", length(time_dim$values))
cat("First 3 steps:", unlist(head(time_dim$values, 3)))
```

The layer spans from 1987 to the present with yearly time steps indexed to January 1st of each year. We now have all four pieces of information needed to download: the **product** ID (`MEDSEA_MULTIYEAR_PHY_006_004`), the **layer** ID (yearly means, `cmems_mod_med_phy-tem_my_4.2km_P1Y-m`), the **variable** ID (`bottomT`), and the available **time range**.

### Choosing a Time Period

To choose a meaningful time range for the temperature data, we should match it to when our benthos occurrences were actually sampled:

```{r}
#| label: fig-occurrence-year-distribution
#| code-fold: true
#| fig-cap: "Number of benthic occurrence records per sampling year"
#| fig-height: 3.5
ggplot(benthos_occurrences, aes(x = year)) +
  geom_bar(aes(fill = after_stat(count)), width = 0.8) +
  scale_fill_continuous(guide = "none", trans = "reverse") +
  scale_x_continuous(breaks = sort(unique(benthos_occurrences$year))) +
  labs(x = "Year", y = "Number of records") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Sampling is concentrated in two periods: a large 1998 campaign and a main cluster from 2004 to 2018 with internal gaps. We download yearly temperature spanning 1998 to 2018, covering both periods.

### Downloading Bottom Temperature

`cms_download_subset()` returns a [`stars`](../concepts.qmd#stars---spatiotemporal-arrays) object (a multi-dimensional array). Note that the `timerange` dates must match the layer's time steps, which as we saw above are indexed to January 1st of each year:

```{r}
#| label: download-bottom-temperature
bottom_temp_stars <- cms_download_subset(
  product = "MEDSEA_MULTIYEAR_PHY_006_004",
  layer = "cmems_mod_med_phy-tem_my_4.2km_P1Y-m",
  variable = "bottomT",
  region = study_bbox,
  timerange = c("1998-01-01", "2018-01-01")
)
```

### Inspecting the Stars Object

Let's examine the structure. Printing a `stars` object shows its attribute (the variable, with summary statistics) and its dimensions:

```{r}
#| label: inspect-stars-object
bottom_temp_stars
```

`st_dimensions()` shows the three dimensions of the object:

```{r}
#| label: inspect-stars-dims
stars::st_dimensions(bottom_temp_stars)
```

Each yearly time step is a 2D spatial grid (longitude × latitude), and the full object stacks these along the time dimension. We can inspect the time values directly:

```{r}
#| label: inspect-stars-time
time_vals <- stars::st_get_dimension_values(bottom_temp_stars, "time")
cat("Time steps:", length(time_vals))
cat("First:", format(time_vals[1]))
cat("Last:", format(time_vals[length(time_vals)]))
```

### Filtering to Sampled Years

Since our download spans 1998-2018 but not all years have occurrence data, we filter the time dimension to keep only years when sampling actually occurred. This gives a temperature mean that reflects the conditions organisms were experiencing.

```{r}
#| label: filter-stars-sampled-years
sampled_years <- unique(benthos_occurrences$year)
keep <- as.integer(format(time_vals, "%Y")) %in% sampled_years
bottom_temp_sampled <- bottom_temp_stars[,,, keep]

cat("Retained", sum(keep), "of", length(keep), "yearly time steps")
cat("Years kept:", sort(unique(as.integer(format(time_vals[keep], "%Y")))))
```

`stars` objects are indexed as `[attribute, dim1, dim2, dim3]`, where the first position selects attributes (variables like `bottomT`) and the rest select along the spatial and temporal dimensions. Leaving a position empty keeps all values along that axis. So `[,,,keep]` means: all attributes, all longitudes, all latitudes, filter time.

### Computing the Mean Temperature

`st_apply()` works like R's `apply()`: the first argument names the dimensions to **keep**, and the function is applied over all remaining dimensions. Here, `c("longitude", "latitude")` keeps the spatial dimensions and applies `mean` across time, collapsing all retained yearly time steps into a single 2D layer:

```{r}
#| label: compute-mean-temperature
bottom_temp_mean <- bottom_temp_sampled |>
  stars::st_apply(c("longitude", "latitude"), mean, na.rm = TRUE)
```

### Converting Stars to Terra Raster

The rest of our analysis uses `terra` for raster operations (zonal statistics, extraction). To convert from `stars` to a `terra` `SpatRaster`, we need to address a grid compatibility issue: the Copernicus Marine grid is **rectilinear** (cell spacing along longitude and latitude is not perfectly uniform), but `terra::rast()` requires a **regular** grid with uniform cell sizes.

`st_warp()` resamples a `stars` object onto a target grid. We build that target with `st_as_stars()`, passing `st_bbox(bottom_temp_mean)` to preserve the original spatial extent and setting `dx` and `dy` to 1/24° to match the native Copernicus Marine resolution (~4.2 km at this latitude):

```{r}
#| label: convert-to-terra
bottom_temp_mean <- bottom_temp_mean |>
  stars::st_warp(stars::st_as_stars(
    sf::st_bbox(bottom_temp_mean),
    dx = 1 / 24,
    dy = 1 / 24
  )) |>
  terra::rast()
names(bottom_temp_mean) <- "bottom_temp_C"

bottom_temp_mean
```

### Visualising Bottom Temperature

```{r}
#| label: fig-bottom-temperature
#| code-fold: true
#| fig-cap: "Mean bottom temperature in the Gulf of Lion (sampled years, 1998-2018)"

ggplot() +
  geom_spatraster(data = bottom_temp_mean) +
  scale_fill_viridis_c(
    name = "Temperature (°C)",
    na.value = "gray90",
    option = "turbo"
  ) +
  labs(title = "Mean Bottom Temperature (sampled years)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

::: {.callout-note collapse="true"}
## What else is available from Copernicus Marine?

In this tutorial we retrieved bottom temperature from a physical reanalysis product. The Copernicus Marine Service also provides:

-   **Other physical variables**: Salinity, currents, sea level, mixed layer depth
-   **Biogeochemical products**: Nutrients, chlorophyll, dissolved oxygen, pH
-   **Multiple temporal resolutions**: Daily, monthly, and climatological means
-   **Multiple depths**: Surface to bottom, including full 3D fields

Browse available products with `cms_products_list()` and follow the same discovery workflow shown above. See the [CopernicusMarine package documentation](https://pepijn-devries.github.io/CopernicusMarine/) for further examples.
:::

## Linking Trait Data from MOBS

The Marine Organism Body Size (MOBS) database provides body size measurements for marine species, linked via WoRMS AphiaIDs.

### About MOBS

The MOBS database [@mccallum2025] compiles body size measurements (length, diameter/width, height) from literature sources for marine species, linked via WoRMS AphiaIDs. This makes it ideal for linking to occurrence data like ours.

### Loading MOBS Data

The MOBS database is available as a CSV file on GitHub, which we can download directly with `readr::read_csv()`. The database uses WoRMS AphiaIDs as its species identifier (allowing us to join to our occurrence data), stores body size measurements (length, diameter/width, height) in centimetres, and conveniently includes taxonomic classification (phylum, class, order, etc.). We pipe the result through `janitor::clean_names()` to convert the mixed-case column names (e.g. `valid_aphiaID`, `scientificName`) to consistent snake_case.

```{r}
#| label: download-mobs
#| cache: true
mobs_url <- "https://raw.githubusercontent.com/crmcclain/MOBS_OPEN/main/data_all_112224.csv"
mobs_raw <- readr::read_csv(mobs_url, show_col_types = FALSE) |>
  janitor::clean_names()

head(mobs_raw, 10)
```

MOBS records body size using different measurements depending on the organism: length, diameter/width, or height. Rather than restricting to a single measurement type (which would lose entire taxonomic groups like Hydrozoa), we take the largest available linear dimension for each record as a general body size descriptor, and convert from centimetres to millimetres. We use `pmax()` (parallel maximum) to do this row-wise: unlike `max()` which returns a single value across all inputs, `pmax()` compares values across columns for each row and returns the largest, ignoring `NA`s. We then filter to species present in our occurrence data:

```{r}
#| label: filter-mobs-to-occurrences
occurrence_aphiaids <- unique(benthos_occurrences$aphiaid)

mobs <- mobs_raw |>
  mutate(
    body_size_mm = pmax(length_cm, diameter_width_cm, height_cm, na.rm = TRUE) *
      10
  ) |>
  filter(valid_aphia_id %in% occurrence_aphiaids, !is.na(body_size_mm)) |>
  select(aphiaid = valid_aphia_id, scientific_name, body_size_mm, phylum, class)

cat("Rows matched:", nrow(mobs))
cat("Unique species:", n_distinct(mobs$aphiaid))
cat("Duplicated AphiaIDs:", sum(duplicated(mobs$aphiaid)))
```

Some species have multiple measurements from different literature sources. For our analysis we need a single value per species, so we take the maximum body size:

```{r}
#| label: summarise-mobs-per-species
mobs <- mobs |>
  group_by(aphiaid, scientific_name, phylum, class) |>
  summarise(max_body_size_mm = max(body_size_mm), .groups = "drop")
```

Since we filtered MOBS to species present in our occurrence data (via `occurrence_aphiaids`), we can check what proportion of our occurrence species have body size data:

```{r}
#| label: mobs-match-rate
cat(sprintf(
  "Matched %d of %d occurrence species (%.1f%%) to MOBS body size data",
  nrow(mobs),
  length(occurrence_aphiaids),
  100 * nrow(mobs) / length(occurrence_aphiaids)
))
```

```{r}
#| label: preview-mobs-matched
head(mobs, 10)
```

### Body Size Distribution

Let's examine the distribution of body sizes among our matched species:

```{r}
#| label: fig-body-size-distribution
#| code-fold: true
#| fig-cap: "Distribution of maximum body size for Gulf of Lion benthic species with MOBS data (log scale)"

n_bins <- 30
log_range <- diff(range(log10(mobs$max_body_size_mm)))
binwidth <- log_range / n_bins

ggplot(mobs, aes(x = max_body_size_mm)) +
  geom_histogram(
    bins = n_bins,
    fill = "#0077BE",
    colour = "white",
    alpha = 0.7
  ) +
  geom_density(
    aes(
      y = after_stat(density) * nrow(mobs) * binwidth,
      colour = "Kernel density estimate"
    ),
    linewidth = 0.8,
    key_glyph = "path"
  ) +
  scale_colour_manual(values = "black", name = NULL) +
  scale_x_log10() +
  labs(
    title = "Body Size Distribution",
    subtitle = sprintf(
      "n = %d Gulf of Lion species with MOBS data",
      nrow(mobs)
    ),
    x = "Maximum body size (mm, log scale)",
    y = "Number of species"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

The distribution is roughly log-normal with a median around 30 mm, but the density curve reveals clear bimodality with peaks near 10 mm and 50 mm. Breaking this down by taxonomic class (@fig-bodysize-by-class) helps explain the pattern.

```{r}
#| label: fig-bodysize-by-class
#| code-fold: true
#| fig-cap: "Body size distributions by taxonomic class for Gulf of Lion benthic species with MOBS length data (classes with fewer than 5 species excluded)"
#| fig-height: 4

class_counts <- mobs |>
  filter(!is.na(class)) |>
  count(class) |>
  filter(n >= 5) |>
  mutate(label = sprintf("%s (n = %d)", class, n))

plot_data <- mobs |>
  inner_join(class_counts, by = "class") |>
  mutate(label = reorder(label, max_body_size_mm, FUN = median))

ggplot(plot_data, aes(x = label, y = max_body_size_mm)) +
  geom_violin(fill = "#0077BE", alpha = 0.5, colour = "white") +
  geom_boxplot(width = 0.15, outlier.size = 0.8, alpha = 0.5) +
  scale_y_log10() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Maximum body size (mm, log scale)"
  ) +
  theme_minimal()
```

The first peak in the histogram is driven by malacostracan crustaceans (`r sum(plot_data$class == "Malacostraca")` species), while the second reflects the larger polychaetes (`r sum(plot_data$class == "Polychaeta")` species) and bivalves (`r sum(plot_data$class == "Bivalvia")` species). Together these three classes make up `r round(100 * sum(plot_data$class %in% c("Malacostraca", "Polychaeta", "Bivalvia")) / nrow(mobs))`% of the matched species.

## Spatial Data Integration

Now we bring our data sources together. First, we enrich `seabed_habitats` with environmental summaries (mean depth and temperature) for each polygon. Then we create `benthos_joined` by using `sf::st_join()` to spatially assign each species record to the habitat polygon it falls within, giving it that polygon's substrate and biozone, and joining body size data from MOBS via AphiaID.

### Checking CRS Alignment

Before combining spatial datasets, we verify their [coordinate reference systems](../concepts.qmd#coordinate-reference-systems-crs). Both vector datasets were downloaded in EPSG:3035, while the rasters are in WGS 84 (EPSG:4326):

```{r}
#| label: check-crs-alignment
# Vector datasets (EPSG:3035)
st_crs(seabed_habitats)$input
st_crs(benthos_occurrences)$input

# Raster datasets (WGS 84)
crs(depth, describe = TRUE)$name
crs(bottom_temp_mean, describe = TRUE)$name
```

The vector datasets share EPSG:3035, so spatial joins will work directly. For the raster extractions below, we reproject the rasters to match the polygons.

### Extracting Environmental Conditions

Rather than sampling a single centroid point per polygon, we compute **zonal statistics**, summarising all raster cells that fall within each habitat polygon. `terra::extract()` takes a raster and a set of polygons, identifies which raster cells fall within each polygon, and applies a summary function (here `mean`) across them. This captures the spatial extent of each habitat and produces more robust environmental characterisations, especially for large or irregularly shaped polygons. Depth and temperature have different native resolutions, so we extract each separately to avoid resampling.

Spatial operations like calculating distances, areas, and zonal statistics require a projected coordinate system with linear units (metres), not geographic coordinates (degrees). In a geographic CRS like WGS 84, the ground distance represented by one degree of longitude varies with latitude, so calculations based on cell counts or coordinate differences will be distorted. An [equal-area projection](../concepts.qmd#why-this-matters-for-marine-spatial-analysis) like EPSG:3035, chosen to suit our European study area, ensures that every raster cell covers the same ground area, so each contributes equally to the mean. We reproject the rasters to match our habitat polygons before extraction:

```{r}
#| label: extract-habitat-environment
seabed_habitats <- seabed_habitats |>
  mutate(
    mean_depth_m = terra::extract(
      terra::project(depth, "EPSG:3035"),
      seabed_habitats,
      fun = mean,
      na.rm = TRUE
    )$depth_m,
    mean_temp_c = terra::extract(
      terra::project(bottom_temp_mean, "EPSG:3035"),
      seabed_habitats,
      fun = mean,
      na.rm = TRUE
    )$bottom_temp_C
  )

names(seabed_habitats)
```

### Joining Occurrences to Habitat Polygons and Traits

We combine the body size join (by AphiaID) and the spatial habitat polygon join (by geographic location) in a single pipeline.

The first step uses `inner_join()`, a standard `dplyr` join that matches rows by shared values in a column (`aphiaid`). Unlike `left_join()` which keeps all rows from the left table, `inner_join()` only retains rows where the key exists in **both** tables. This means only occurrence records whose species also appear in MOBS (i.e. have body size data) are kept.

Although the occurrence data already contains pre-matched EUNIS codes, we use a **spatial join** rather than joining on those codes. This is an important distinction: a join on EUNIS codes would associate each occurrence with every polygon sharing that habitat type, duplicating records across polygons where the species was never actually recorded. A spatial join instead links each occurrence to the specific polygon it falls within, preserving the actual environmental and habitat conditions in which it was recorded.

We perform spatial joins with `sf::st_join()` which works differently from a standard `dplyr` join: instead of matching values in columns, it matches on **geometry**. For each point in the left dataset (occurrences), `st_join()` finds which polygon in the right dataset (habitat polygons) contains that point, and appends that polygon's attributes. The default spatial predicate is `st_intersects`, meaning a point is matched to any polygon it touches or falls inside. This is the spatial equivalent of `left_join()`: all rows from the left dataset are kept, and unmatched rows get `NA` for the joined columns. Because both datasets are in EPSG:3035, the spatial join uses planar geometry directly (no geographic coordinate warnings).

We `select()` only `substrate` and `biozone` from the habitat polygons before joining, so only those columns are appended. Finally, we filter out any occurrences without a biozone value, keeping only records we can assign to a biozone for our characterisation:

```{r}
#| label: join-occurrences-habitats-traits
benthos_joined <- benthos_occurrences |>
  inner_join(mobs, by = "aphiaid") |>
  st_join(select(seabed_habitats, substrate, biozone)) |>
  filter(!is.na(biozone))

cat(sprintf(
  "Retained %d of %d occurrences (%.1f%%) with body size and habitat data",
  nrow(benthos_joined),
  nrow(benthos_occurrences),
  100 * nrow(benthos_joined) / nrow(benthos_occurrences)
))
```

A preview of the joined dataset, showing each occurrence with its species name, body size, EUNIS classification, substrate, and biozone:

```{r}
#| label: preview-joined-occurrences
benthos_joined |>
  st_drop_geometry() |>
  select(scientificname, max_body_size_mm, EUNIS2019C, substrate, biozone) |>
  head(10)
```

## Characterising Biozones

Biozones represent the vertical zonation of the seabed, from the sunlit infralittoral through the circalittoral to the deep bathyal and abyssal zones. This depth gradient is the primary ecological axis structuring benthic communities, influencing light availability, temperature, pressure, and food supply. The EUSeaMap habitat layer encodes biozone alongside habitat type and substrate for each polygon, giving us a natural framework to bring together all our data sources.

### Habitat Composition by Biozone

Different EUNIS habitat types are distributed across biozones. Using the full set of habitat polygons in the study area, this stacked bar chart shows which habitat types occur in each zone:

```{r}
#| label: fig-habitat-by-biozone
#| code-fold: true
#| fig-cap: "Proportional EUNIS habitat type composition within each biozone"
#| fig-height: 9

habitat_biozone_data <- seabed_habitats |>
  filter(!is.na(biozone), !is.na(eunis2019d)) |>
  count(biozone, eunis2019d)

ggplot(
  habitat_biozone_data,
  aes(x = biozone, y = n, fill = str_wrap(eunis2019d, width = 50))
) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_viridis_d(option = "turbo", name = NULL) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Biozone", y = "Proportion of polygons") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.key.size = unit(0.4, "cm")
  ) +
  guides(fill = guide_legend(ncol = 2, override.aes = list(alpha = 1)))
```

Habitat diversity varies across biozones. The infralittoral supports the greatest variety of EUNIS habitat types (including Posidonia beds, rock, sand, and coarse sediment). The shallow circalittoral retains several types (coralligène, coarse and mixed sediments, muddy detritic bottoms), while the deep circalittoral is dominated by a single detritic habitat. The bathyal zones show renewed diversity with rock, coarse sediment, mixed sediment, sand, and mud all represented, though mud becomes increasingly dominant. Only the abyssal maps onto a single habitat type (Mediterranean abyssal mud).

### Substrate Composition by Biozone

Substrate type (the physical material of the seabed) also varies with depth:

```{r}
#| label: fig-substrate-by-biozone
#| code-fold: true
#| fig-cap: "Proportional substrate composition within each biozone"
#| fig-height: 5

substrate_biozone_data <- seabed_habitats |>
  filter(!is.na(biozone), !is.na(substrate)) |>
  count(biozone, substrate)

ggplot(substrate_biozone_data, aes(x = biozone, y = n, fill = substrate)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_brewer(palette = "Set3", name = "Substrate") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Biozone", y = "Proportion of polygons") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Substrate composition also varies with depth. The infralittoral is the most diverse, featuring Posidonia meadows, coarse and mixed sediment, rock, sand, and dead mattes. Sand dominates the shallow circalittoral, giving way to muddy sand and fine mud in the deep circalittoral. The bathyal zones retain a mix of substrates (rock, muddy sand, sand, fine mud) with fine mud becoming increasingly prominent. The abyssal zone is entirely fine mud.

### Environmental Conditions by Biozone

Using the depth (EMODnet WCS) and bottom temperature (Copernicus Marine) data we extracted earlier, we can characterise the environmental conditions within each biozone. Each point represents one habitat polygon:

```{r}
#| label: fig-env-by-biozone
#| code-fold: true
#| fig-cap: "Depth and bottom temperature distributions across biozones"
#| fig-height: 4
#| fig-width: 10

env_data <- seabed_habitats |>
  filter(!is.na(biozone), !is.na(mean_depth_m))

p_depth <- ggplot(env_data, aes(x = biozone, y = mean_depth_m)) +
  geom_violin(fill = "#0077BE", alpha = 0.5, colour = "white") +
  geom_boxplot(width = 0.15, outlier.size = 0.8, alpha = 0.5) +
  labs(x = NULL, y = "Mean depth (m)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_temp <- ggplot(
  env_data |> filter(!is.na(mean_temp_c)),
  aes(x = biozone, y = mean_temp_c)
) +
  geom_violin(fill = "#E74C3C", alpha = 0.5, colour = "white") +
  geom_boxplot(width = 0.15, outlier.size = 0.8, alpha = 0.5) +
  labs(x = NULL, y = "Mean bottom temperature (\u00b0C)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_depth + p_temp
```

Depth increases steadily across biozones as expected, from near-surface in the infralittoral to over 2,500 m in the abyssal. The upper and lower bathyal zones show the greatest variability in depth, spanning several hundred metres each. Bottom temperature shows a narrower overall range (~13 to 17 °C): the infralittoral and shallow circalittoral are warmer and more variable (reflecting seasonal and coastal influences), while the deeper biozones converge on a narrow band around 13 °C.

### Body Size Distributions by Biozone

Finally, we examine how body size varies across biozones. Ridge plots (from [`ggridges`](https://wilkelab.org/ggridges/)) show kernel density estimates stacked vertically, with fill colour mapped to body size. `scale` controls how much adjacent ridges overlap (0.9 keeps them mostly separate), and `rel_min_height` trims density tails below 1% of the peak to avoid long, thin artefacts:

```{r}
#| label: fig-bodysize-by-biozone
#| code-fold: true
#| fig-cap: "Body size distribution by biozone"
#| fig-height: 5

biozone_plot_data <- benthos_joined |>
  st_drop_geometry() |>
  filter(!is.na(biozone), !is.na(max_body_size_mm)) |>
  group_by(biozone) |>
  filter(n() >= 10) |>
  ungroup()

ggplot(
  biozone_plot_data,
  aes(x = max_body_size_mm, y = factor(biozone, levels = rev(levels(biozone))))
) +
  geom_density_ridges_gradient(
    aes(fill = after_stat(x)),
    scale = 0.9,
    rel_min_height = 0.01
  ) +
  scale_x_log10() +
  scale_fill_viridis_c(
    name = "Body size (mm)",
    option = "plasma",
    trans = "log10",
    labels = scales::label_comma()
  ) +
  labs(
    x = "Maximum body size (mm, log scale)",
    y = "Biozone"
  ) +
  theme_minimal()
```

Biozones with fewer than 10 records are excluded from the plot. The upper bathyal has too few records with body size data, and no occurrences fall within the abyssal zone. Among the remaining biozones, the lower bathyal is dominated by smaller-bodied species, while the shallower zones (infralittoral, circalittoral) show broader body size distributions extending to larger species. This pattern is consistent with the shift in taxonomic composition and environmental conditions across depth zones.

We can break this down further by looking at taxonomic class within each biozone, using the same violin + boxplot approach as @fig-bodysize-by-class:

```{r}
#| label: fig-bodysize-class-biozone
#| code-fold: true
#| fig-cap: "Body size distributions by taxonomic class within each biozone (classes with fewer than 5 species excluded)"
#| fig-height: 8
#| fig-width: 10

class_biozone_data <- benthos_joined |>
  st_drop_geometry() |>
  filter(!is.na(biozone), !is.na(max_body_size_mm), !is.na(class)) |>
  group_by(biozone) |>
  filter(n() >= 10) |>
  ungroup()

# Keep only classes with >= 5 species across all biozones
class_keep <- class_biozone_data |>
  distinct(aphiaid, class) |>
  count(class) |>
  filter(n >= 5)

class_biozone_data <- class_biozone_data |>
  semi_join(class_keep, by = "class") |>
  mutate(class = reorder(class, max_body_size_mm, FUN = median))

ggplot(class_biozone_data, aes(x = class, y = max_body_size_mm)) +
  geom_violin(fill = "#0077BE", alpha = 0.5, colour = "white") +
  geom_boxplot(width = 0.15, outlier.size = 0.5, alpha = 0.5) +
  scale_y_log10() +
  coord_flip() +
  facet_wrap(~biozone) +
  labs(
    x = NULL,
    y = "Maximum body size (mm, log scale)"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

Breaking body size down by taxonomic class within each biozone reveals which groups drive the overall patterns. The shallower biozones (infralittoral and shallow circalittoral) support the widest range of classes and body sizes. Malacostraca consistently appear as the smallest-bodied group across all biozones. The deeper biozones retain fewer classes, with polychaetes, bivalves, and malacostraca persisting into the deep circalittoral and lower bathyal while groups like anthozoa and echinoidea drop out.

## Interactive Map

The plots above characterise each biozone in isolation. An interactive map brings these layers together spatially, letting you see where habitats, species occurrences, and body size classes are distributed across the Gulf of Lion. You can zoom into shelf features and submarine canyons, click individual occurrences for species details, and toggle habitat polygons on and off to explore spatial relationships that static plots cannot capture.

In earlier tutorials we used `tmap` for interactive maps. Here we use the [`leaflet`](https://rstudio.github.io/leaflet/) package, which provides a direct R interface to the [Leaflet.js](https://leafletjs.com/) library. While `tmap` is excellent for quick thematic maps, `leaflet` gives finer control over layer styling, pop-up content, and rendering behaviour, making it a valuable tool to have in your spatial analysis toolkit.

Unlike `tmap`, `leaflet` requires all data in WGS 84 (EPSG:4326), so both layers need reprojecting with `st_transform()`. The habitat polygons also contain very detailed coastline and boundary geometry that is unnecessary for an overview map and slows down interactive rendering, so we additionally repair any invalid geometries with `st_make_valid()` (common in complex polygon datasets from web services) and reduce vertex counts with `st_simplify()`. The `dTolerance` parameter sets the maximum deviation in CRS units (here metres for EPSG:3035):

```{r}
#| label: prepare-habitats-leaflet
habitats_leaflet <- seabed_habitats |>
  st_make_valid() |>
  st_simplify(dTolerance = 200) |>
  st_transform(4326)
```

For the occurrence points, we bin continuous body sizes into discrete size classes using `cut()`, creating categories that are easy to distinguish on a colour scale, and reproject to WGS 84:

```{r}
#| label: prepare-occurrences-leaflet
benthos_leaflet <- benthos_joined |>
  filter(!is.na(max_body_size_mm)) |>
  mutate(
    size_class = cut(
      max_body_size_mm,
      breaks = c(0, 10, 50, 200, 1000, Inf),
      labels = c("<10mm", "10-50mm", "50-200mm", "200-1000mm", ">1000mm"),
      right = FALSE
    )
  ) |>
  st_transform(4326)
```

We build the map in layers. `addPolygons()` draws the habitat polygons, using `colorFactor()` to map EUNIS habitat codes to colours from the `"turbo"` palette. `addCircleMarkers()` overlays occurrence points coloured by body size class. Both layers include clickable pop-ups with relevant attributes:

```{r}
#| label: fig-interactive-map
#| eval: false

# Colour palettes
habitat_pal <- colorFactor("turbo", domain = habitats_leaflet$eunis2019c)
size_pal <- colorFactor("viridis", domain = benthos_leaflet$size_class)

leaflet() |>
  addProviderTiles("Esri.WorldGrayCanvas") |>
  addPolygons(
    data = habitats_leaflet,
    fillColor = ~ habitat_pal(eunis2019c),
    fillOpacity = 0.5,
    weight = 0,
    smoothFactor = 0,
    group = "Habitats",
    popup = ~ paste0(
      "<b>",
      eunis2019d,
      "</b><br>",
      "<b>Biozone:</b> ",
      biozone,
      "<br>",
      "<b>Substrate:</b> ",
      substrate,
      "<br>",
      "<b>Mean depth (m):</b> ",
      round(mean_depth_m, 1),
      "<br>",
      "<b>Mean bottom temp (\u00B0C):</b> ",
      round(mean_temp_c, 1)
    )
  ) |>
  addCircleMarkers(
    data = benthos_leaflet,
    radius = 4,
    fillColor = ~ size_pal(size_class),
    fillOpacity = 0.7,
    weight = 0.5,
    color = "white",
    group = "Occurrences",
    popup = ~ paste0(
      "<b>Species:</b> ",
      scientificname,
      "<br>",
      "<b>Body size (mm):</b> ",
      max_body_size_mm,
      "<br>",
      "<b>Habitat:</b> ",
      EUNIS2019C,
      "<br>",
      "<b>Biozone:</b> ",
      biozone
    )
  ) |>
  addLegend(
    pal = size_pal,
    values = benthos_leaflet$size_class,
    title = "Body size",
    group = "Occurrences"
  ) |>
  addLayersControl(
    overlayGroups = c("Habitats", "Occurrences"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

```{r}
#| label: fig-interactive-map-render
#| echo: false
#| fig-cap: "Interactive map of benthic species occurrences coloured by body size class, overlaid on seabed habitat polygons. Click on features for details."

# Colour palettes
habitat_pal <- colorFactor("turbo", domain = habitats_leaflet$eunis2019c)
size_pal <- colorFactor("viridis", domain = benthos_leaflet$size_class)

interactive_map <- leaflet() |>
  addProviderTiles("Esri.WorldGrayCanvas") |>
  addPolygons(
    data = habitats_leaflet,
    fillColor = ~ habitat_pal(eunis2019c),
    fillOpacity = 0.5,
    weight = 0,
    smoothFactor = 0,
    group = "Habitats",
    popup = ~ paste0(
      "<b>",
      eunis2019d,
      "</b><br>",
      "<b>Biozone:</b> ",
      biozone,
      "<br>",
      "<b>Substrate:</b> ",
      substrate,
      "<br>",
      "<b>Mean depth (m):</b> ",
      round(mean_depth_m, 1),
      "<br>",
      "<b>Mean bottom temp (\u00B0C):</b> ",
      round(mean_temp_c, 1)
    )
  ) |>
  addCircleMarkers(
    data = benthos_leaflet,
    radius = 4,
    fillColor = ~ size_pal(size_class),
    fillOpacity = 0.7,
    weight = 0.5,
    color = "white",
    group = "Occurrences",
    popup = ~ paste0(
      "<b>Species:</b> ",
      scientificname,
      "<br>",
      "<b>Body size (mm):</b> ",
      max_body_size_mm,
      "<br>",
      "<b>Habitat:</b> ",
      EUNIS2019C,
      "<br>",
      "<b>Biozone:</b> ",
      biozone
    )
  ) |>
  addLegend(
    pal = size_pal,
    values = benthos_leaflet$size_class,
    title = "Body size",
    group = "Occurrences"
  ) |>
  addLayersControl(
    overlayGroups = c("Habitats", "Occurrences"),
    options = layersControlOptions(collapsed = FALSE)
  )

# Save to standalone HTML to avoid pandoc hanging on large htmlwidgets
htmlwidgets::saveWidget(
  interactive_map,
  get_output_path("biozone_bodysize_map.html", "tutorials"),
  selfcontained = TRUE
)

knitr::include_url("biozone_bodysize_map.html", height = "600px")
```

The `popup` arguments use HTML strings to format the pop-up content. Tags like `<b>...</b>` for bold text and `<br>` for line breaks give control over how information is presented when a user clicks a feature. This is a key advantage of `leaflet` over `tmap`: full HTML support in pop-ups means you can style text, add links, or even embed images. See the [Leaflet for R documentation on popups and labels](https://rstudio.github.io/leaflet/articles/popups.html) for more examples.

Note the `smoothFactor = 0` argument in `addPolygons()`. By default, Leaflet applies its own Douglas-Peucker simplification to polygon coordinates at render time to improve performance. For complex or concave polygons this can cause the simplified outline to cross itself, producing visible line artifacts across the polygon fill. Since we have already simplified the geometry server-side with `st_simplify()`, setting `smoothFactor = 0` disables this additional client-side step and avoids these artifacts.

## Discussion

### Key Findings

This analysis demonstrates the integration of multiple EU marine data infrastructures to characterise benthic biozones in the Gulf of Lion:

1.  **Habitat composition varies across biozones**: Shallow zones are dominated by sandy and coarse substrates, while deeper zones shift towards muddy sediments
2.  **Environmental gradients structure biozones**: Depth and temperature covary with biozone, confirming that biozones capture meaningful environmental variation
3.  **Body size shifts with depth**: Shallower biozones tend to support species with larger body sizes, with a shift towards smaller-bodied fauna in deeper zones
4.  **Data integration**: Linking EMODnet occurrence and habitat data with Copernicus environmental layers and external trait databases enables multi-faceted ecological characterisation

### Ecological Interpretation

Biozones represent the vertical zonation of the seabed, reflecting gradients in light, pressure, temperature, and food supply. Our results are consistent with several ecological expectations:

-   **Depth zonation**: The infralittoral and circalittoral zones, receiving more light and organic input, support diverse habitat types and substrates
-   **Body size patterns**: The tendency for smaller body sizes at depth may reflect reduced food availability in deeper zones, consistent with Bergmann's rule inversions observed in some marine taxa
-   **Substrate control**: Substrate composition changes with depth and hydrodynamic regime, shaping habitat availability

However, these interpretations require caution:

1.  **Sampling bias**: Occurrence records may not represent true community composition, and shallow zones are typically better sampled
2.  **Taxonomic coverage**: MOBS trait data is incomplete, with smaller and less-studied taxa underrepresented
3.  **Spatial resolution**: Environmental data resolution may not capture fine-scale habitat heterogeneity

### Limitations

-   **Trait coverage**: Not all species in the occurrence data have body size records in MOBS, so body size patterns reflect a subset of the community
-   **Temporal mismatch**: Occurrence records span different time periods than environmental data
-   **Single region**: Patterns observed in the Gulf of Lion may not generalise to other Mediterranean basins with different oceanographic settings

## What You've Learned

This tutorial demonstrated advanced multi-source data integration:

-   **EMODnet WFS**: Accessing seabed habitat polygons and species occurrence records
-   **EMODnet WCS**: Retrieving bathymetry rasters for zonal statistics
-   **Copernicus Marine Service**: Downloading bottom temperature data
-   **External databases**: Linking to body size trait data (MOBS) via taxonomic identifiers
-   **Spatial operations**: Zonal statistics, CRS reprojection, and spatial joins across data types
-   **Biozone characterisation**: Using biozones as an organising framework to synthesise habitat, environmental, and trait data

These skills enable complex marine ecological analyses that draw on the full breadth of available EU marine data.

## Further Resources

-   [EMODnet Seabed Habitats Portal](https://emodnet.ec.europa.eu/en/seabed-habitats): Habitat maps
-   [EMODnet Biology Portal](https://emodnet.ec.europa.eu/en/biology): Species occurrence data
-   [Copernicus Marine Service](https://marine.copernicus.eu/): Oceanographic data
-   [World Register of Marine Species (WoRMS)](https://www.marinespecies.org/): Taxonomic backbone
-   [MOBS Database](https://github.com/crmcclain/MOBS_OPEN): Marine body size data

## References
