---
title: "4: Characterising Benthic Biozones"
subtitle: "Combining Species Occurrences, Habitat Maps, Bathymetry, Temperature, and Body Size Data in the Gulf of Lion"
difficulty: "Advanced"
time: "~90 minutes"
data-type: "WFS + WCS + External sources"
description: "Learn to integrate EMODnet WFS and WCS services with external data sources including the Copernicus Marine Service and trait databases. Characterise benthic biozones in the Gulf of Lion using habitat composition, substrate type, environmental conditions, and body size distributions."
datasets: "Seabed habitat polygons, benthic species occurrences, bathymetry, bottom temperature, body size data"
df-print: paged
bibliography: references.bib
format:
  html:
    embed-resources: false
    code-link: true
---

## Introduction

The Gulf of Lion, located in the northwestern Mediterranean off the coast of southern France, hosts a rich diversity of benthic communities shaped by the Rhône River delta, diverse substrates from sandy beaches to deep canyons, and strong environmental gradients. Characterising the biozones of this region, from the sunlit infralittoral to the deep bathyal, requires bringing together habitat maps, environmental data, species occurrences, and functional traits such as body size.

In this tutorial, we demonstrate how to **integrate multiple EU marine data infrastructures** to characterise benthic biozones in the Gulf of Lion. We combine:

-   **EMODnet WFS** for habitat maps and species occurrence records
-   **EMODnet WCS** for bathymetry data
-   **Copernicus Marine Service** for bottom temperature
-   **MOBS database** for species body size data

This multi-source approach reflects real-world marine ecological analysis, where insights emerge from combining data across platforms and domains.

### Why Include Body Size?

Body size is a fundamental trait linked to metabolism, feeding strategy, life history, and ecological role. Adding body size to a biozone characterisation enriches the picture beyond habitat composition and physical conditions. The distribution of body sizes across biozones can reveal:

-   **Environmental filtering**: Temperature and depth gradients may favour certain body size ranges
-   **Habitat associations**: Different seabed types may support communities with distinct size structures
-   **Community function**: Body size distributions affect energy flow and ecosystem processes

### Learning Objectives

By the end of this tutorial, you will be able to:

-   Retrieve seabed habitat polygons from EMODnet Seabed Habitats WFS
-   Access benthic species occurrence data from EMODnet Biology WFS
-   Download bathymetry rasters from EMODnet Bathymetry WCS
-   Access bottom temperature data from the Copernicus Marine Service
-   Link species records to body size data from external databases
-   Join datasets using shared identifiers (EUNIS codes, AphiaIDs)
-   Extract zonal statistics from rasters using habitat polygons
-   Characterise benthic biozones using habitat, environmental, and trait data

### Data Sources

**EMODnet WFS (Vector Data):**

| Layer | Service | Description |
|------------------|----------------------|--------------------------------|
| `eusm2025_eunis2019_full` | seabed_habitats_general_datasets_and_products | EU Seabed Map with EUNIS 2019 classification |
| `eb185_benthos_full_matched` | biology | Benthic species occurrence records from European seas |

**EMODnet WCS (Raster Data):**

| Coverage             | Service    | Description                   |
|----------------------|------------|-------------------------------|
| `emodnet__mean_2022` | bathymetry | Mean water depth (bathymetry) |

**External Sources:**

| Source | Description |
|----------------------------|--------------------------------------------|
| [Copernicus Marine Service](https://marine.copernicus.eu/) | Bottom temperature from [MEDSEA_MULTIYEAR_PHY_006_004](https://data.marine.copernicus.eu/product/MEDSEA_MULTIYEAR_PHY_006_004/description) reanalysis |
| [MOBS Database](https://github.com/crmcclain/MOBS_OPEN) | Marine Organism Body Size database [@mccallum2025] |

## Setup

### Packages

This tutorial uses a broader set of packages than previous tutorials, reflecting the multi-source integration workflow:

```{r}
#| label: load-packages
#| message: false
library(emodnet.wfs)
library(emodnet.wcs)
library(CopernicusMarine)
library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(ggridges)
library(patchwork)
library(tidyterra)
library(tmap)
```

```{r}
#| label: load-internal-packages
#| include: false
library(emodnet.bio.r.geo.tutorials)
```

::: {.callout-tip collapse="true"}
## Package installation

Most packages are available from CRAN:

``` r
install.packages(c(
  "emodnet.wfs", "CopernicusMarine", "terra", "sf", "stars",
  "dplyr", "tidyr", "purrr", "readr", "stringr", "ggplot2",
  "ggridges", "janitor", "patchwork", "tidyterra", "tmap",
  "rnaturalearth"
))
```

The `emodnet.wcs` package is not yet on CRAN; install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("EMODnet/emodnet.wcs")
```
:::

### Study Area

We focus on the Gulf of Lion in the northwestern Mediterranean. This region offers good coverage across most data sources involved, with diverse substrate types (sand, mud, coarse sediments, seagrass meadows) and spatial sampling across 165 unique locations. The Gulf of Lion also spans multiple biozones from infralittoral to bathyal depths:

```{r}
#| label: define-study-area-bbox
# Study area: Gulf of Lion for habitats and environmental data
study_bbox <- c(xmin = 3, ymin = 42.5, xmax = 7, ymax = 44)

# Create bbox polygon for visualisation (WGS 84) and spatial operations (EPSG:3035)
bbox_polygon <- sf::st_as_sfc(sf::st_bbox(study_bbox, crs = 4326))
bbox_polygon_3035 <- sf::st_transform(bbox_polygon, 3035)
```

```{r}
#| label: fig-study-area
#| code-fold: true
#| fig-cap: "Study area: Gulf of Lion in the northwestern Mediterranean"

# Get Mediterranean coastlines
med_countries <- rnaturalearth::ne_countries(
  scale = "medium",
  returnclass = "sf"
) |>
  sf::st_crop(c(xmin = -2, ymin = 38, xmax = 12, ymax = 46))

ggplot() +
  geom_sf(
    data = med_countries,
    fill = "gray90",
    color = "gray60",
    linewidth = 0.3
  ) +
  geom_sf(data = bbox_polygon, fill = NA, color = "#004494", linewidth = 1.5) +
  coord_sf(xlim = c(-2, 12), ylim = c(38, 46)) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(panel.grid = element_line(color = "gray85", linewidth = 0.2))
```

## Retrieving Seabed Habitat Data

Our first data source is seabed habitat polygons from EMODnet Seabed Habitats. These polygons define the spatial framework for our biozone characterisation, providing EUNIS habitat codes, substrate types, and biozone classifications across the study area.

### Connecting to EMODnet Seabed Habitats WFS

For an introduction to working with WFS services and the `emodnet.wfs` package, see [Tutorial 1](tutorial-01.qmd). Here we apply the same workflow to the Seabed Habitats service.

```{r}
#| label: init-habitats-wfs
habitat_wfs <- emodnet_init_wfs_client(
  service = "seabed_habitats_general_datasets_and_products"
)
```

### Exploring Habitat Layers

We need habitat polygons from the [EUSeaMap](https://emodnet.ec.europa.eu/en/euseamap), the broad-scale predictive habitat map for European seas. The service contains many layers, so we filter by title to find the EUSeaMap products:

```{r}
#| label: explore-habitat-layers
#| cache: true
habitat_layers <- emodnet_get_wfs_info(habitat_wfs)

# Search for EUSeaMap layers by title
habitat_layers |>
  filter(grepl("EUSeaMap", title)) |>
  select(layer_name, title)
```

The results show separate layers for energy, substrate, biozone, and two habitat classification systems (EUNIS 2007 and EUNIS 2019). We choose `eusm2025_eunis2019_full` because EUNIS 2019 is the current classification standard and this layer includes both the habitat code and the substrate and biozone attributes we need.

### Understanding Habitat Classifications

Let's examine the attribute structure of the habitat layer:

```{r}
#| label: inspect-habitat-attributes
#| cache: true
habitat_attrs <- layer_attribute_descriptions(
  wfs = habitat_wfs,
  layer = "eusm2025_eunis2019_full"
)

habitat_attrs
```

Key attributes include:

-   **`eunis2019c`**: EUNIS 2019 habitat classification (level 3)
-   **`eunis2019d`**: Human-readable EUNIS 2019 habitat description
-   **`substrate`**: Seabed substrate type
-   **`biozone`**: Biological zone

Let's check what habitat values are present across the full layer (not just our study area) using `layer_attribute_inspect()`:

```{r}
#| label: explore-habitat-types
#| cache: true
eunis_values <- layer_attribute_inspect(
  wfs = habitat_wfs,
  layer = "eusm2025_eunis2019_full",
  attribute = "eunis2019c"
)

eunis_values
```

Scrolling through the values, you will notice that not all entries are clean EUNIS codes. Some contain `"Na"` (polygons with no habitat classification) and others contain `"or"` (e.g. `"MB551 or MB552"`), representing polygons where the classification is ambiguous between two habitat types. Let's quantify how many polygons across the entire layer are affected:

```{r}
#| label: count-invalid-habitat-values
eunis_values |>
  filter(.data[["."]] == "Na" | grepl("or", .data[["."]]))
```

These are data gaps in the EUSeaMap product itself, not specific to our study area. Since these values are not valid EUNIS codes, they would cause problems in downstream joins. We filter them out server-side using CQL to avoid downloading polygons we cannot use.

### Building an Attribute CQL Filter

We start by extracting the geometry column name for the BBOX filter:

```{r}
#| label: habitat-geom-col
habitat_geom_col <- habitat_attrs$name[habitat_attrs$geometry]
```

We also want to combine the spatial BBOX filter with **attribute filters** to exclude the problematic `eunis2019c` values we identified above. CQL supports combining conditions with `AND`: we use `<>` to exclude exact matches and `NOT LIKE` with `%` wildcards for pattern matching. For more on attribute filtering syntax, see the [emodnet.wfs ECQL filtering vignette](https://emodnet.github.io/emodnet.wfs/articles/ecql_filtering.html#filtering-by-attributes).

```{r}
#| label: habitat-cql-filter
habitat_cql_filter <- sprintf(
  "BBOX(%s, %s, %s, %s, %s, 'EPSG:4326') AND eunis2019c <> 'Na' AND eunis2019c NOT LIKE '%%or%%'",
  habitat_geom_col,
  study_bbox["xmin"],
  study_bbox["ymin"],
  study_bbox["xmax"],
  study_bbox["ymax"]
)

habitat_cql_filter
```

The `<>` operator excludes polygons where `eunis2019c` is exactly `"Na"`, and `NOT LIKE '%or%'` excludes any value containing the substring `"or"` (the `%` characters are CQL wildcards matching any sequence of characters). Note the `%%` in the `sprintf()` call: this is R's escape sequence to produce a literal `%` in the output string.

### Downloading Habitat Polygons

Now we download with the combined spatial and attribute filter. The habitat layer's native CRS is Web Mercator (EPSG:3857), but we request **EPSG:3035** (ETRS89-extended / LAEA Europe) via `crs = 3035`. This is an [equal-area projection](../concepts.qmd#why-this-matters-for-marine-spatial-analysis) well suited to European marine data: every cell and polygon covers the same ground area regardless of latitude, which matters for zonal statistics and spatial joins later in the tutorial:

```{r}
#| label: download-habitat-polygons
#| cache: true
seabed_habitats <- emodnet_get_layers(
  wfs = habitat_wfs,
  layers = "eusm2025_eunis2019_full",
  cql_filter = habitat_cql_filter,
  outputFormat = "application/json",
  simplify = TRUE,
  crs = 3035
)

nrow(seabed_habitats)
```

### Cropping Polygons to the Study Area

The WFS server returns any polygon that overlaps our bounding box, but it returns each polygon in its entirety. Some polygons, particularly large deep-sea habitat types like the abyssal plain, may extend far beyond our study area. We can see this by checking the bounding box of the downloaded data:

```{r}
#| label: check-habitat-bbox
st_bbox(seabed_habitats)
```

The extent is much larger than our study area (`r study_bbox["xmin"]`--`r study_bbox["xmax"]`°E, `r study_bbox["ymin"]`--`r study_bbox["ymax"]`°N). Leaving these oversized polygons in place causes two problems: it can skew analyses by including data outside the area of interest, and later, when we extract zonal statistics from rasters constrained to our bounding box, the overhanging parts of polygons produce `NaN` values (the raster has no cells to average over those areas).

We fix this with `sf::st_crop()`, which clips all geometries to a target rectangle. We use `bbox_polygon_3035`, the EPSG:3035 version of our study area polygon created at the start of the tutorial:

```{r}
#| label: crop-habitat-polygons
seabed_habitats <- seabed_habitats |>
  st_crop(bbox_polygon_3035)

st_bbox(seabed_habitats)
```

The bounding box now matches our study area.

### Inspecting Downloaded Data

Let's preview the key columns. Each row is a polygon with a EUNIS habitat code, a human-readable description, substrate type, and biozone:

```{r}
#| label: preview-habitat-polygons
seabed_habitats |>
  select(eunis2019c, eunis2019d, substrate, biozone) |>
  head(10)
```

::: {.callout-note}
The WFS layer may use a different name for its geometry column (e.g. `the_geom`, `msGeometry`), but after download the `sf` package standardises the column name to `geometry`. This is normal and applies to all WFS downloads throughout this tutorial.
:::

The geometry column contains `sf MULTIPOLYGON` features. Each habitat feature is a polygon (or collection of polygons) delineating an area of seabed with a particular classification. The columns show:

- **`eunis2019c`**: the EUNIS 2019 level 3 code (e.g. `MB552`)
- **`eunis2019d`**: the human-readable habitat description
- **`substrate`**: the physical substrate (e.g. Sand, Mud, Coarse substrate)
- **`biozone`**: the depth zone (e.g. Infralittoral, Deep circalittoral)

The `biozone` variable represents depth zones from shallow (Infralittoral) to deep (Abyssal). We convert it to an ordered factor for meaningful analysis along depth gradients, and standardise `substrate` casing (the source data mixes "SAND" and "Sand"):

```{r}
#| label: order-biozone
biozone_levels <- c(
  "Infralittoral",
  "Shallow circalittoral",
  "Deep circalittoral",
  "Upper bathyal",
  "Lower bathyal",
  "Abyssal"
)

seabed_habitats <- seabed_habitats |>
  mutate(
    biozone = factor(biozone, levels = biozone_levels, ordered = TRUE),
    substrate = str_to_sentence(substrate)
  )
```

We can confirm the CQL filter worked by checking the unique `eunis2019c` values in the downloaded data:

```{r}
#| label: verify-clean-eunis
sort(unique(seabed_habitats$eunis2019c))
```

All clean EUNIS codes, with no `"Na"` or `"or"` values. Let's see which habitat types are most common in our study area:

```{r}
#| label: habitat-type-summary
# Summarise habitat types present
seabed_habitats |>
  st_drop_geometry() |>
  count(eunis2019c, sort = TRUE) |>
  head(10)
```

### Visualising Seabed Habitats

```{r}
#| label: fig-seabed-habitats
#| code-fold: true
#| fig-cap: "Seabed habitat classification in the Gulf of Lion (EUNIS 2019)"
#| fig-height: 10
#| fig-width: 8

ggplot() +
  geom_sf(data = med_countries, fill = "gray90", colour = "gray70") +
  geom_sf(
    data = seabed_habitats,
    aes(fill = str_wrap(eunis2019d, width = 50)),
    colour = NA,
    alpha = 0.8
  ) +
  coord_sf(
    xlim = c(study_bbox["xmin"], study_bbox["xmax"]),
    ylim = c(study_bbox["ymin"], study_bbox["ymax"])
  ) +
  scale_fill_viridis_d(option = "turbo", name = NULL) +
  labs(title = "Seabed Habitats (EUNIS 2019)") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.key.size = unit(0.4, "cm"),
    plot.margin = margin(2, 2, 2, 2)
  ) +
  guides(fill = guide_legend(ncol = 2, override.aes = list(alpha = 1)))
```

## Retrieving Species Occurrence Data

Next, we retrieve benthic species occurrence records from EMODnet Biology. These records have been matched to the World Register of Marine Species (WoRMS), ensuring taxonomic consistency, and each record is pre-matched to a EUNIS habitat code via the EUSeaMap grid, allowing us to link occurrences to the habitat polygons we just downloaded.

::: {.callout-tip collapse="true"}
## Taxonomic matching with the worrms package

The data we use here are already matched to WoRMS, but if you need to match your own species lists to the WoRMS taxonomic backbone, the [`worrms`](https://cran.r-project.org/package=worrms) R package provides a programmatic interface to WoRMS. It supports fuzzy name matching, retrieval of accepted names and synonyms, and lookup of taxonomic hierarchies, useful for standardising species names across datasets.
:::

### Connecting to EMODnet Biology WFS

We follow the same discovery workflow as for the habitat service above.

```{r}
#| label: init-biology-wfs
bio_wfs <- emodnet_init_wfs_client(service = "biology")
```

### Exploring Available Layers

Let's search for benthic occurrence data:

```{r}
#| label: explore-biology-layers
#| cache: true
bio_layers <- emodnet_get_wfs_info(bio_wfs)

# Filter for benthos-related layers
bio_layers |>
  filter(grepl("benthos", layer_name, ignore.case = TRUE)) |>
  select(layer_name, title, abstract) |>
  # Keep first paragraph of abstract for display
  mutate(abstract = str_extract(abstract, "^[^\n]+")) |>
  knitr::kable()
```

We choose `eb185_benthos_full_matched` because it contains individual occurrence records with coordinates, each matched to WoRMS (providing AphiaIDs we can use to link to trait data) and pre-matched to EUNIS habitat codes via the EUSeaMap grid. The `benthos_europe` layer, by contrast, provides gridded presence/absence summaries rather than individual occurrence records, so it is less suited to point-level analysis.

### Understanding Layer Structure

Before downloading, let's examine what attributes are available (see [Tutorial 1](tutorial-01.qmd#understanding-layer-structure) for more on inspecting WFS layer metadata):

```{r}
#| label: inspect-benthos-attributes
#| cache: true
benthos_attrs <- layer_attribute_descriptions(
  wfs = bio_wfs,
  layer = "eb185_benthos_full_matched"
)

benthos_attrs
```

Key attributes for our analysis include:

-   **`st_point`**: Geometry column. We need to know this column name to construct spatial CQL filters (e.g. bounding box queries), so we extract it programmatically below
-   **`aphiaid`**: WoRMS identifier for linking to trait databases
-   **`scientificname`**: Species name for display
-   **`lon`**, **`lat`**: Coordinates
-   **`year`**: Sampling year
-   **`eusm_polygon_id`**: Link to the EUSeaMap grid polygon
-   **`EUNIS2019C`**, **`EUNIS2019D`**: EUNIS 2019 habitat classification (level 3 and 4), already matched via the EUSeaMap grid

### Downloading Occurrence Data

We use a [CQL BBOX filter](tutorial-01.qmd#filtering-by-location-pre-download) to request only records within our study area:

```{r}
#| label: define-benthos-bbox-filter
# Get the geometry column name
benthos_geom_col <- benthos_attrs$name[benthos_attrs$geometry]

# Create ECQL BBOX filter for study area
benthos_bbox_filter <- sprintf(
  "BBOX(%s, %s, %s, %s, %s, 'EPSG:4326')",
  benthos_geom_col,
  study_bbox["xmin"],
  study_bbox["ymin"],
  study_bbox["xmax"],
  study_bbox["ymax"]
)
```

The resulting CQL filter combines the geometry column name with the `study_bbox` coordinates into a string that the WFS server interprets as a spatial constraint:

```{r}
#| label: print-benthos-bbox-filter
benthos_bbox_filter
```

Now we download the data with `emodnet_get_layers()`, passing our CQL filter to restrict the query to the study area. As with the habitat data, we request EPSG:3035 so both vector datasets share the same equal-area CRS. We use [GeoJSON format](tutorial-01.qmd#why-use-geojson-format) (`outputFormat = "application/json"`) for faster downloads and simpler geometry types, and set `simplify = TRUE` to return a single `sf` data frame rather than a list of layers:

```{r}
#| label: download-benthos-occurrences
#| cache: true
benthos_occurrences <- emodnet_get_layers(
  wfs = bio_wfs,
  layers = "eb185_benthos_full_matched",
  cql_filter = benthos_bbox_filter,
  outputFormat = "application/json",
  simplify = TRUE,
  crs = 3035
)

# Check the result
nrow(benthos_occurrences)
```

### Inspecting Downloaded Data

A quick preview of the data. Note the `sf POINT` geometry type in the geometry column: each occurrence is a single coordinate, in contrast to the habitat **polygons** we retrieved earlier:

```{r}
#| label: preview-benthos
benthos_occurrences |>
  select(aphiaid, scientificname, lon, lat, year, EUNIS2019C) |>
  head(10)
```

Each occurrence has been pre-matched to a EUNIS 2019 habitat code. Let's see which habitat types are represented and confirm the codes are clean:

```{r}
#| label: preview-benthos-eunis
sort(unique(benthos_occurrences$EUNIS2019C))
```

Let's examine the taxonomic diversity and data completeness:

```{r}
#| label: benthos-summary
# How many unique species?
n_species <- n_distinct(benthos_occurrences$aphiaid)

# How many records have valid AphiaIDs for trait matching?
n_with_aphiaid <- sum(!is.na(benthos_occurrences$aphiaid))

cat(sprintf(
  "Total records: %d\nUnique species (AphiaIDs): %d\nRecords with valid AphiaID: %d (%.1f%%)\n",
  nrow(benthos_occurrences),
  n_species,
  n_with_aphiaid,
  100 * n_with_aphiaid / nrow(benthos_occurrences)
))
```

### Quick Visualisation

```{r}
#| label: fig-benthos-occurrences
#| code-fold: true
#| fig-cap: "Benthic species occurrence records in the Gulf of Lion"

ggplot() +
  geom_sf(data = med_countries, fill = "gray90", colour = "gray70") +
  geom_sf(
    data = benthos_occurrences,
    colour = "#0077BE",
    size = 0.5,
    alpha = 0.3
  ) +
  coord_sf(
    xlim = c(study_bbox["xmin"], study_bbox["xmax"]),
    ylim = c(study_bbox["ymin"], study_bbox["ymax"])
  ) +
  labs(
    title = "Benthic Species Occurrences",
    subtitle = sprintf(
      "%d records in the Gulf of Lion",
      nrow(benthos_occurrences)
    )
  ) +
  theme_minimal()
```

## Retrieving Bathymetry Data

Bathymetry (water depth) provides essential environmental context for our habitat characterisation. We retrieve this from EMODnet Bathymetry using [WCS](../concepts.qmd#wcs-web-coverage-service) (see [Tutorial 2](tutorial-02.qmd) for an introduction to working with the `emodnet.wcs` package).

### Connecting to EMODnet Bathymetry WCS

```{r}
#| label: init-bathymetry-wcs
bathy_wcs <- emdn_init_wcs_client(service = "bathymetry")
```

### Exploring Available Coverages

`emdn_get_coverage_ids()` lists all available coverage layers from the service:

```{r}
#| label: list-bathymetry-coverages
bathy_coverage_ids <- emdn_get_coverage_ids(bathy_wcs)
bathy_coverage_ids
```

The coverage IDs encode the product type and reference year. The `mean` coverages provide the gridded mean depth, with year suffixes indicating different releases of the bathymetric compilation. We use the most recent release (`emodnet__mean_2022`). `emdn_get_coverage_info()` provides metadata including the spatial resolution, extent, and coordinate reference system:

```{r}
#| label: get-bathymetry-info
bathy_info <- emdn_get_coverage_info(
  wcs = bathy_wcs,
  coverage_ids = "emodnet__mean_2022"
)

bathy_info |>
  knitr::kable()
```

The resolution is approximately 1/960 of a degree (~115 m at this latitude), which is much finer than our bottom temperature raster (~4.2 km). This difference in resolution is one reason we extract zonal statistics from each raster separately later on.

### Downloading Bathymetry

We download the bathymetry raster for our study area using `emdn_get_coverage()`, which returns a `SpatRaster`. We set `nil_values_as_na = TRUE` so that any nil values declared in the coverage metadata (e.g. `-9999` for "no data") are converted to R's `NA`:

```{r}
#| label: download-bathymetry
bathymetry <- emdn_get_coverage(
  wcs = bathy_wcs,
  coverage_id = "emodnet__mean_2022",
  bbox = study_bbox,
  nil_values_as_na = TRUE
)

bathymetry
```

### Inspecting Downloaded Data

EMODnet bathymetry uses the convention of **negative values for depth below sea level** and positive values for elevation above it. Since our bounding box overlaps the coastline, the raster includes both marine cells (negative) and land cells (positive). We can check the value range with `terra::global()`, which computes a summary statistic across all cells of a `SpatRaster`:

```{r}
#| label: check-bathymetry-range
terra::global(bathymetry, range, na.rm = TRUE)
```

The negative minimum confirms deep marine areas. The positive maximum (~1,900 m) may seem surprising for a marine study area, but our bounding box extends to 44°N, which reaches well inland over southern France into the foothills of the Pyrenees and Massif Central. EMODnet bathymetry is a digital terrain model that covers both sea and land, so these terrestrial elevations are included in the download. We set land cells (positive values) to `NA`, then negate the remaining values so that depth is expressed as a positive number (deeper = larger):

```{r}
#| label: process-bathymetry
bathymetry[bathymetry > 0] <- NA
depth <- -1 * bathymetry
names(depth) <- "depth_m"
```

### Visualising Bathymetry

```{r}
#| label: fig-bathymetry
#| code-fold: true
#| fig-cap: "Bathymetry of the Gulf of Lion (depth in metres)"

ggplot() +
  geom_spatraster(data = depth) +
  geom_sf(data = med_countries, fill = "gray40", colour = "gray30") +
  scale_fill_viridis_c(
    name = "Depth (m)",
    na.value = "white",
    option = "mako",
    direction = -1,
    breaks = c(0, 200, 500, 1000, 2000, 2500),
    # Allocate more of the colour gradient to shallow depths so that
    # shelf features and canyon heads are visible despite the large
    # overall depth range (~0–2700 m)
    values = scales::rescale(
      c(0, 50, 200, 500, 1000, 2000, 2500),
      to = c(0, 1),
      from = c(0, 2500)
    )
  ) +
  coord_sf(
    xlim = c(study_bbox["xmin"], study_bbox["xmax"]),
    ylim = c(study_bbox["ymin"], study_bbox["ymax"])
  ) +
  labs(title = "Gulf of Lion Bathymetry") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(
    fill = guide_colorbar(
      barwidth = 15,
      barheight = 0.5,
      title.position = "top"
    )
  )
```

## Retrieving Bottom Temperature

Bottom temperature provides insight into thermal niches and is a key variable for characterising environmental conditions across biozones. We access this from the [Copernicus Marine Service](https://marine.copernicus.eu/) (CMEMS), the marine component of the EU's [Copernicus](https://www.copernicus.eu/) Earth observation programme. Copernicus is the Earth observation component of the EU's Space programme, drawing on satellite and in-situ observations to monitor the planet. All Copernicus data and products are provided on a free, full and open basis under EU regulation. CMEMS provides ocean analysis, forecast and reanalysis products covering the global ocean and European regional seas. We interact with it in R using the [`CopernicusMarine`](https://pepijn-devries.github.io/CopernicusMarine/) package.

### How Copernicus Marine Service data is organised

The EMODnet services we used above follow [OGC standards](../concepts.qmd#ogc-standards-and-spatial-data-types): WFS serves vector features and WCS serves raster coverages, each accessed through a service client. Copernicus Marine Service organises its data differently. Data is stored in both NetCDF and [zarr](../concepts.qmd#zarr-.zarr) (a cloud-native array format) and discovered through a [STAC](https://stacspec.org/) (SpatioTemporal Asset Catalog) catalogue. The `CopernicusMarine` package handles this transparently (using the zarr stores for efficient spatial subsetting), but the terminology differs from the OGC services:

| EMODnet | Copernicus Marine | Copernicus Example |
|---------|-------------------|--------------------|
| Service (e.g., `"biology"`) | **Product** (a dataset with a unique ID) | `MEDSEA_MULTIYEAR_PHY_006_004` |
| Layer (e.g., `"eb185_benthos_full_matched"`) | **Layer** (a specific temporal resolution or subset) | `cmems_mod_med_phy-temp_my_4.2km_P1M-m` |
| Attributes (WFS) or Bands (WCS) | **Variables** (physical quantities stored in the array) | `bottomT`, `thetao` |
| 2D features or grids | **Multi-dimensional arrays** with longitude, latitude, depth, and time | 4D datacube |

Because the data is multi-dimensional, `cms_download_subset()` returns a [`stars`](../concepts.qmd#stars---spatiotemporal-arrays) object rather than the `sf` or `terra` objects returned by EMODnet services. For more background, see the [stars](../concepts.qmd#stars---spatiotemporal-arrays) and [zarr](../concepts.qmd#zarr-.zarr) sections in the Geospatial Concepts page.

The metadata returned by `CopernicusMarine` functions follows the STAC specification. For a general introduction to what STAC is and how catalogues, collections, and items relate, see the [STAC intro tutorial](https://stacspec.org/en/tutorials/intro-to-stac/). In practice, the deeply nested metadata structures can be challenging to navigate, but contain rich information about layers, variables, dimensions, and time ranges, as we explore below.

The discovery workflow follows a similar pattern to EMODnet: list what is available, inspect metadata, then download a subset.

### Choosing a Data Product

Bottom temperature is typically available in physical reanalysis products, identified by `MULTIYEAR_PHY` in their product IDs. `cms_products_list()` lists all available products; we filter to physical reanalysis:

```{r}
#| label: explore-copernicus-products
#| cache: true
# List physical reanalysis products
cms_products_list() |>
  filter(grepl("MULTIYEAR_PHY", product_id)) |>
  select(product_id, title) |>
  knitr::kable()
```

A Copernicus **product** plays the same role as an EMODnet **service** (like `"biology"` or `"bathymetry"`): it is a thematic collection that groups related layers together. For small study areas, regional products typically offer higher resolution and are specifically validated for that region. For our Gulf of Lion study, the Mediterranean product ([MEDSEA_MULTIYEAR_PHY_006_004](https://data.marine.copernicus.eu/product/MEDSEA_MULTIYEAR_PHY_006_004/description)) provides \~4.2 km resolution compared to \~8 km for the global product.

### Browsing Available Variables

Unlike OGC services where each layer contains a single dataset, Copernicus products are built on a datacube architecture. A single product can bundle multiple physical variables (temperature, salinity, currents) across multiple layers. We can get a high-level overview of all variables available across the entire product using `cms_product_details()`:

```{r}
#| label: explore-med-product-details
#| cache: true
med_product <- cms_product_details("MEDSEA_MULTIYEAR_PHY_006_004")

# allVariables aggregates human-readable variable names across ALL layers
med_product$properties$allVariables |> unlist()
```

This confirms the product includes "Sea water potential temperature at sea floor", which is what we need. However, this is a product-wide summary: it aggregates variable names across all layers, so not every variable listed here is necessarily available in every layer. To find out exactly which variables are available in which layers, and to get the variable `id`s needed for downloading, we need to drill into the layer-level metadata.

### Discovering Layers and Their Variables

A product contains multiple **layers**, each offering a different temporal resolution of the same underlying data (e.g. daily vs monthly means). `cms_product_services()` lists them:

```{r}
#| label: explore-med-services
#| cache: true
med_services <- cms_product_services("MEDSEA_MULTIYEAR_PHY_006_004")

# Each row is a layer
med_services$id
```

The layer `id`s encode the variable group and temporal resolution (e.g., `temp` for temperature, `P1M-m` for monthly means, `P1D-m` for daily). Version suffixes (e.g., `_202511`) can be omitted when downloading.

Each layer carries rich variable metadata in its `cube:variables` property. The [STAC datacube extension](https://github.com/stac-extensions/datacube) defines a base set of fields (`dimensions`, `type`, `unit`, `description`), and Copernicus Marine adds its own fields on top. Let's examine the structure by drilling into the first variable of the first layer. `med_services` is a data frame with list-columns, where each row is a layer. `properties[[1]]` accesses the STAC metadata for the first layer, `cube:variables` is a named list of that layer's variables, and `[[1]]` selects the first variable entry:

```{r}
#| label: inspect-cube-variables
# Examine the metadata structure of the first variable in the first layer
str(med_services$properties[[1]]$`cube:variables`[[1]], max.level = 2)
```

The standard STAC datacube fields include `dimensions` (which axes the variable spans), `type` (`"data"` for measured values), and `unit`. In standard STAC, each variable is identified by its key name in the `cube:variables` map. Copernicus Marine also stores this as an explicit `id` field inside each variable object, along with other Copernicus-specific additions: `name$en` (a human-readable description), `standardName` (the [CF convention](https://cfconventions.org/Data/cf-standard-names/current/build/cf-standard-name-table.html) name), and visualisation hints like `colormapId` and value ranges.

We can use these fields to build a clear mapping of layers to their variables. Since we need bottom temperature, we filter to the temperature layers:

```{r}
#| label: check-temp-layer-variables
# Extract variable details from each temperature layer
med_services |>
  filter(grepl("tem", id, ignore.case = TRUE)) |>
  rowwise() |>
  reframe(
    layer = id,
    variable = purrr::map_chr(properties$`cube:variables`, "id"),
    description = purrr::map_chr(properties$`cube:variables`, \(x) x$name$en),
    unit = purrr::map_chr(properties$`cube:variables`, "unit")
  ) |>
  knitr::kable()
```

All temperature layers contain `bottomT` (sea floor potential temperature) and `thetao` (potential temperature through the water column). For comprehensive documentation of all variable names, layer naming conventions, and production methodology, see the [Product User Manual (PUM)](https://documentation.marine.copernicus.eu/PUM/CMEMS-MED-PUM-006-004.pdf).

We choose the monthly means layer (`P1M-m`) rather than daily (`P1D-m`) or climatology. Monthly data provides enough temporal resolution to compute a meaningful multi-year average while keeping the download size manageable. The climatology layer would seem ideal (it provides pre-computed long-term averages), but it has additional dimensions that the `CopernicusMarine` package cannot currently handle. Computing our own average from monthly data gives us full control over the time period.

```{r}
#| label: select-temp-layer
# Select the monthly temperature layer
temp_monthly <- med_services |>
  filter(id == "cmems_mod_med_phy-temp_my_4.2km_P1M-m_202511")
```

The `cube:dimensions` property tells us the available **temporal extent** and **time steps** for our chosen layer, which is essential for knowing what date range we can request:

```{r}
#| label: check-temp-layer-time
time_dim <- temp_monthly$properties[[1]]$`cube:dimensions`$time

# Temporal extent (start and end dates)
cat("Temporal extent:", time_dim$extent[[1]], "to", time_dim$extent[[2]])
cat("Total time steps:", length(time_dim$values))
```

The layer spans from 1987 to the present with monthly time steps. We now have all four pieces of information needed to download: the **product** ID, the **layer** ID (we choose monthly means, `P1M-m`), the **variable** ID (`bottomT`), and the available **time range**.

### Choosing a Time Period

To choose a meaningful time range for the temperature data, we should match it to when our benthos occurrences were actually sampled:

```{r}
#| label: fig-occurrence-year-distribution
#| code-fold: true
#| fig-cap: "Number of benthic occurrence records per sampling year"
#| fig-height: 3.5
ggplot(benthos_occurrences, aes(x = year)) +
  geom_bar(aes(fill = after_stat(count)), width = 0.8) +
  scale_fill_continuous(guide = "none", trans = "reverse") +
  scale_x_continuous(breaks = sort(unique(benthos_occurrences$year))) +
  labs(x = "Year", y = "Number of records") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Sampling is concentrated in two periods: a large 1998 campaign and a main cluster from 2004 to 2018 with internal gaps. We download monthly temperature spanning 1998 to 2015, covering both periods while keeping the download small (~6 MB).

### Downloading Bottom Temperature

`cms_download_subset()` returns a [`stars`](../concepts.qmd#stars---spatiotemporal-arrays) object (a multi-dimensional array):

```{r}
#| label: download-bottom-temperature
bottom_temp_stars <- cms_download_subset(
  product = "MEDSEA_MULTIYEAR_PHY_006_004",
  layer = "cmems_mod_med_phy-temp_my_4.2km_P1M-m",
  variable = "bottomT",
  region = study_bbox,
  timerange = c("1998-01-01", "2015-12-31")
)
```

### Inspecting the Stars Object

Let's examine the structure. Printing a `stars` object shows its attribute (the variable, with summary statistics) and its dimensions:

```{r}
#| label: inspect-stars-object
bottom_temp_stars
```

`st_dimensions()` shows the three dimensions of the object:

```{r}
#| label: inspect-stars-dims
stars::st_dimensions(bottom_temp_stars)
```

Each monthly time step is a 2D spatial grid (longitude × latitude), and the full object stacks these along the time dimension. We can inspect the time values directly:

```{r}
#| label: inspect-stars-time
time_vals <- stars::st_get_dimension_values(bottom_temp_stars, "time")
cat("Time steps:", length(time_vals))
cat("First:", format(time_vals[1]))
cat("Last:", format(time_vals[length(time_vals)]))
```

### Filtering to Sampled Years

Since our download spans 1998-2015 but not all years have occurrence data, we filter the time dimension to keep only months from years when sampling actually occurred. This gives a temperature mean that reflects the conditions organisms were experiencing.

```{r}
#| label: filter-stars-sampled-years
sampled_years <- unique(benthos_occurrences$year)
keep <- as.integer(format(time_vals, "%Y")) %in% sampled_years
bottom_temp_sampled <- bottom_temp_stars[,,, keep]

cat("Retained", sum(keep), "of", length(keep), "monthly time steps")
cat("Years kept:", sort(unique(as.integer(format(time_vals[keep], "%Y")))))
```

`stars` objects are indexed as `[attribute, dim1, dim2, dim3]`, where the first position selects attributes (variables like `bottomT`) and the rest select along the spatial and temporal dimensions. Leaving a position empty keeps all values along that axis. So `[,,,keep]` means: all attributes, all longitudes, all latitudes, filter time.

### Computing the Mean Temperature

`st_apply()` works like R's `apply()`: the first argument names the dimensions to **keep**, and the function is applied over all remaining dimensions. Here, `c("longitude", "latitude")` keeps the spatial dimensions and applies `mean` across time, collapsing all retained monthly time steps into a single 2D layer:

```{r}
#| label: compute-mean-temperature
bottom_temp_mean <- bottom_temp_sampled |>
  stars::st_apply(c("longitude", "latitude"), mean, na.rm = TRUE)
```

### Converting Stars to Terra Raster

The rest of our analysis uses `terra` for raster operations (zonal statistics, extraction). To convert from `stars` to a `terra` `SpatRaster`, we need to address a grid compatibility issue: the Copernicus Marine grid is **rectilinear** (cell spacing along longitude and latitude is not perfectly uniform), but `terra::rast()` requires a **regular** grid with uniform cell sizes.

`st_warp()` resamples a `stars` object onto a target grid. We build that target with `st_as_stars()`, passing `st_bbox(bottom_temp_mean)` to preserve the original spatial extent and setting `dx` and `dy` to 1/24° to match the native Copernicus Marine resolution (~4.2 km at this latitude):

```{r}
#| label: convert-to-terra
bottom_temp_mean <- bottom_temp_mean |>
  stars::st_warp(stars::st_as_stars(
    sf::st_bbox(bottom_temp_mean),
    dx = 1 / 24,
    dy = 1 / 24
  )) |>
  terra::rast()
names(bottom_temp_mean) <- "bottom_temp_C"

bottom_temp_mean
```

### Visualising Bottom Temperature

```{r}
#| label: fig-bottom-temperature
#| code-fold: true
#| fig-cap: "Mean bottom temperature in the Gulf of Lion (sampled years, 1998-2015)"

ggplot() +
  geom_spatraster(data = bottom_temp_mean) +
  geom_sf(data = med_countries, fill = "gray40", colour = "gray30") +
  scale_fill_viridis_c(
    name = "Temperature (°C)",
    na.value = "white",
    option = "turbo"
  ) +
  coord_sf(
    xlim = c(study_bbox["xmin"], study_bbox["xmax"]),
    ylim = c(study_bbox["ymin"], study_bbox["ymax"])
  ) +
  labs(title = "Mean Bottom Temperature (sampled years)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

::: {.callout-note collapse="true"}
## What else is available from Copernicus Marine?

In this tutorial we retrieved bottom temperature from a physical reanalysis product. The Copernicus Marine Service also provides:

-   **Other physical variables**: Salinity, currents, sea level, mixed layer depth
-   **Biogeochemical products**: Nutrients, chlorophyll, dissolved oxygen, pH
-   **Multiple temporal resolutions**: Daily, monthly, and climatological means
-   **Multiple depths**: Surface to bottom, including full 3D fields

Browse available products with `cms_products_list()` and follow the same discovery workflow shown above. See the [CopernicusMarine package documentation](https://pepijn-devries.github.io/CopernicusMarine/) for further examples.
:::

## Linking Trait Data from MOBS

The Marine Organism Body Size (MOBS) database provides body size measurements for marine species, linked via WoRMS AphiaIDs.

### About MOBS

The MOBS database [@mccallum2025] compiles body size data from literature sources, standardised to maximum body length where possible. This makes it ideal for linking to occurrence data that includes WoRMS identifiers (AphiaIDs).

### Loading MOBS Data

The MOBS database is available as a CSV file on GitHub, which we can download directly with `readr::read_csv()`. The database uses WoRMS AphiaIDs as its species identifier (allowing us to join to our occurrence data) and stores body size measurements (length, diameter/width, height) in centimetres. We pipe the result through `janitor::clean_names()` to convert the mixed-case column names (e.g. `valid_aphiaID`, `scientificName`) to consistent snake_case.

```{r}
#| label: download-mobs
#| cache: true
mobs_url <- "https://raw.githubusercontent.com/crmcclain/MOBS_OPEN/main/data_all_112224.csv"
mobs_raw <- readr::read_csv(mobs_url, show_col_types = FALSE) |>
  janitor::clean_names()

head(mobs_raw, 10)
```

We'll use the length measurement and convert from centimetres to millimetres. First, we filter MOBS to species present in our occurrence data, keeping only rows that have a length measurement (not all records do), convert units and select the columns we need:

```{r}
#| label: filter-mobs-to-occurrences
occurrence_aphiaids <- unique(benthos_occurrences$aphiaid)

mobs_filtered <- mobs_raw |>
  filter(valid_aphia_id %in% occurrence_aphiaids, !is.na(length_cm)) |>
  mutate(length_mm = length_cm * 10) |>
  select(aphiaid = valid_aphia_id, scientific_name, length_mm, phylum, class)

cat("Rows matched:", nrow(mobs_filtered))
cat("Unique species:", n_distinct(mobs_filtered$aphiaid))
cat("Duplicated AphiaIDs:", sum(duplicated(mobs_filtered$aphiaid)))
```

Some species have multiple measurements from different literature sources. For our analysis we need a single value per species, so we take the maximum body length:

```{r}
#| label: summarise-mobs-per-species
mobs_matched <- mobs_filtered |>
  group_by(aphiaid, scientific_name, phylum, class) |>
  summarise(max_length_mm = max(length_mm), .groups = "drop")

n_matched <- nrow(mobs_matched)
n_total <- length(occurrence_aphiaids)

cat(sprintf(
  "Matched %d of %d species (%.1f%%) to MOBS body size data",
  n_matched,
  n_total,
  100 * n_matched / n_total
))
```

```{r}
#| label: preview-mobs-matched
head(mobs_matched, 10)
```

### Body Size Distribution

Let's examine the distribution of body sizes among our matched species:

```{r}
#| label: fig-body-size-distribution
#| code-fold: true
#| fig-cap: "Distribution of maximum body length for Gulf of Lion benthic species with MOBS data (log scale)"

n_bins <- 30
log_range <- diff(range(log10(mobs_matched$max_length_mm)))
binwidth <- log_range / n_bins

ggplot(mobs_matched, aes(x = max_length_mm)) +
  geom_histogram(
    bins = n_bins,
    fill = "#0077BE",
    colour = "white",
    alpha = 0.7
  ) +
  geom_density(
    aes(
      y = after_stat(density) * nrow(mobs_matched) * binwidth,
      colour = "Kernel density estimate"
    ),
    linewidth = 0.8,
    key_glyph = "path"
  ) +
  scale_colour_manual(values = "black", name = NULL) +
  scale_x_log10() +
  labs(
    title = "Body Size Distribution",
    subtitle = sprintf(
      "n = %d Gulf of Lion species with MOBS data",
      nrow(mobs_matched)
    ),
    x = "Maximum body length (mm, log scale)",
    y = "Number of species"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

The distribution is roughly log-normal with a median around 30 mm, but the density curve reveals clear bimodality with peaks near 10 mm and 50 mm. Breaking this down by taxonomic class (@fig-bodysize-by-class) helps explain the pattern.

```{r}
#| label: fig-bodysize-by-class
#| code-fold: true
#| fig-cap: "Body size distributions by taxonomic class for Gulf of Lion benthic species with MOBS length data (classes with fewer than 5 species excluded)"
#| fig-height: 4

class_counts <- mobs_matched |>
  filter(!is.na(class)) |>
  count(class) |>
  filter(n >= 5) |>
  mutate(label = sprintf("%s (n = %d)", class, n))

plot_data <- mobs_matched |>
  inner_join(class_counts, by = "class") |>
  mutate(label = reorder(label, max_length_mm, FUN = median))

ggplot(plot_data, aes(x = label, y = max_length_mm)) +
  geom_violin(fill = "#0077BE", alpha = 0.5, colour = "white") +
  geom_boxplot(width = 0.15, outlier.size = 0.8) +
  scale_y_log10() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Maximum body length (mm, log scale)"
  ) +
  theme_minimal()
```

The first peak in the histogram is driven by malacostracan crustaceans (190 species, median ~10 mm), while the second reflects the larger polychaetes (231 species) and bivalves (132 species). Together these three classes make up over 90% of the matched species.

## Spatial Data Integration

Now we bring our data sources together. The integration produces two datasets, each serving a different role in the biozone characterisation:

-   **Polygon level** (`habitat_env`): We compute environmental summaries (mean depth and temperature) for each habitat polygon, using the full set of polygons in the study area. This dataset characterises the biozones themselves: their habitat composition, substrate makeup, and environmental conditions.
-   **Occurrence level** (`occurrences_in_habitat`): We use `sf::st_join()` to spatially assign each species record to the habitat polygon it falls within, giving it that polygon's substrate and biozone. We also join body size data from MOBS via AphiaID. This dataset drives the body size analysis.

### Checking CRS Alignment

Before combining spatial datasets, we verify their [coordinate reference systems](../concepts.qmd#coordinate-reference-systems-crs). Both vector datasets were downloaded in EPSG:3035, while the rasters are in WGS 84 (EPSG:4326):

```{r}
#| label: check-crs-alignment
# Vector datasets (EPSG:3035)
st_crs(seabed_habitats)$input
st_crs(benthos_occurrences)$input

# Raster datasets (WGS 84)
crs(depth, describe = TRUE)$name
crs(bottom_temp_mean, describe = TRUE)$name
```

The vector datasets share EPSG:3035, so spatial joins will work directly. For the raster extractions below, we reproject the rasters to match the polygons.

### Extracting Environmental Conditions

Rather than sampling a single centroid point per polygon, we compute **zonal statistics**, summarising all raster cells that fall within each habitat polygon. `terra::extract()` takes a raster and a set of polygons, identifies which raster cells fall within each polygon, and applies a summary function (here `mean`) across them. This captures the spatial extent of each habitat and produces more robust environmental characterisations, especially for large or irregularly shaped polygons. Depth and temperature have different native resolutions, so we extract each separately to avoid resampling.

Because our habitat polygons are in EPSG:3035 (an [equal-area projection](../concepts.qmd#why-this-matters-for-marine-spatial-analysis)), we reproject the rasters to match before extraction. In an equal-area projection every raster cell covers the same ground area regardless of latitude, so each cell contributes equally to the mean:

```{r}
#| label: extract-habitat-environment
habitat_env <- seabed_habitats |>
  st_drop_geometry() |>
  mutate(
    mean_depth_m = terra::extract(
      terra::project(depth, "EPSG:3035"),
      seabed_habitats,
      fun = mean,
      na.rm = TRUE
    )$depth_m,
    mean_temp_C = terra::extract(
      terra::project(bottom_temp_mean, "EPSG:3035"),
      seabed_habitats,
      fun = mean,
      na.rm = TRUE
    )$bottom_temp_C
  )
```

### Joining Occurrences to Habitats and Traits

We combine the body size join (by AphiaID) and the spatial habitat join (by location) in a single pipeline. `inner_join()` retains only species with MOBS body size data, and `st_join()` assigns each occurrence to the habitat polygon it falls within, adding that polygon's `substrate` and `biozone`. Because both datasets are in EPSG:3035, the spatial join uses planar geometry directly:

```{r}
#| label: join-occurrences-habitats-traits
occurrences_in_habitat <- benthos_occurrences |>
  inner_join(mobs_matched, by = "aphiaid") |>
  st_join(select(seabed_habitats, substrate, biozone)) |>
  filter(!is.na(biozone))

cat(sprintf(
  "Retained %d of %d occurrences (%.1f%%) with body size and habitat data",
  nrow(occurrences_in_habitat),
  nrow(benthos_occurrences),
  100 * nrow(occurrences_in_habitat) / nrow(benthos_occurrences)
))
```

A preview of the joined dataset, showing each occurrence with its species name, body size, habitat code, substrate, and biozone:

```{r}
#| label: preview-joined-occurrences
occurrences_in_habitat |>
  st_drop_geometry() |>
  select(scientificname, max_length_mm, EUNIS2019C, substrate, biozone) |>
  head(10)
```

## Characterising Biozones

Biozones represent the vertical zonation of the seabed, from the sunlit infralittoral through the circalittoral to the deep bathyal and abyssal zones. This depth gradient is the primary ecological axis structuring benthic communities, influencing light availability, temperature, pressure, and food supply. The EUSeaMap habitat layer encodes biozone alongside habitat type and substrate for each polygon, giving us a natural framework to bring together all our data sources.

### Habitat Composition by Biozone

Different EUNIS habitat types are distributed across biozones. Using the full set of habitat polygons in the study area, this stacked bar chart shows which habitat types occur in each zone:

```{r}
#| label: fig-habitat-by-biozone
#| code-fold: true
#| fig-cap: "Proportional EUNIS habitat type composition within each biozone"
#| fig-height: 7

habitat_biozone_data <- habitat_env |>
  filter(!is.na(biozone), !is.na(eunis2019d)) |>
  count(biozone, eunis2019d)

ggplot(
  habitat_biozone_data,
  aes(x = biozone, y = n, fill = str_wrap(eunis2019d, width = 50))
) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_viridis_d(option = "turbo", name = NULL) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Biozone", y = "Proportion of polygons") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.key.size = unit(0.4, "cm")
  ) +
  guides(fill = guide_legend(ncol = 2, override.aes = list(alpha = 1)))
```

Habitat diversity decreases sharply with depth. The infralittoral supports the greatest variety of EUNIS habitat types (including Posidonia beds, rock, sand, and coarse sediment), and the shallow circalittoral retains several types (coralligène, coarse and mixed sediments, muddy detritic bottoms). From the deep circalittoral downwards, each biozone maps onto a single habitat type in our matched data, all mud-dominated.

### Substrate Composition by Biozone

Substrate type (the physical material of the seabed) also varies with depth:

```{r}
#| label: fig-substrate-by-biozone
#| code-fold: true
#| fig-cap: "Proportional substrate composition within each biozone"
#| fig-height: 5

substrate_biozone_data <- habitat_env |>
  filter(!is.na(biozone), !is.na(substrate)) |>
  count(biozone, substrate)

ggplot(substrate_biozone_data, aes(x = biozone, y = n, fill = substrate)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_brewer(palette = "Set3", name = "Substrate") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Biozone", y = "Proportion of polygons") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Substrate composition also shows a clear transition with depth. The infralittoral is again the most diverse, featuring Posidonia meadows, coarse and mixed sediment, rock, and sand. The shallow and deep circalittoral retain a mix of substrates (sand, muddy sand, coarse sediment, fine mud), but fine mud becomes increasingly dominant. From the upper bathyal downwards, fine mud dominates all biozones, reaching 100% in the abyssal zone.

### Environmental Conditions by Biozone

Using the depth (EMODnet WCS) and bottom temperature (Copernicus Marine) data we extracted earlier, we can characterise the environmental conditions within each biozone. Each point represents one habitat polygon:

```{r}
#| label: fig-env-by-biozone
#| code-fold: true
#| fig-cap: "Depth and bottom temperature distributions across biozones"
#| fig-height: 4
#| fig-width: 10

env_data <- habitat_env |>
  filter(!is.na(biozone), !is.na(mean_depth_m))

p_depth <- ggplot(env_data, aes(x = biozone, y = mean_depth_m)) +
  geom_violin(fill = "#0077BE", alpha = 0.5, colour = "white") +
  geom_boxplot(width = 0.15, outlier.size = 0.8) +
  labs(x = NULL, y = "Mean depth (m)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_temp <- ggplot(
  env_data |> filter(!is.na(mean_temp_C)),
  aes(x = biozone, y = mean_temp_C)
) +
  geom_violin(fill = "#E74C3C", alpha = 0.5, colour = "white") +
  geom_boxplot(width = 0.15, outlier.size = 0.8) +
  labs(x = NULL, y = "Mean bottom temperature (\u00b0C)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_depth + p_temp
```

Depth increases steadily across biozones as expected, from near-surface in the infralittoral to over 2,500 m in the abyssal. The upper and lower bathyal zones show the greatest variability in depth, spanning several hundred metres each. Bottom temperature shows a narrower overall range (~13 to 17 °C): the infralittoral and shallow circalittoral are warmer and more variable (reflecting seasonal and coastal influences), while the deeper biozones converge on a narrow band around 13 °C.

### Body Size Distributions by Biozone

Finally, we examine how body size varies across biozones. Ridge plots (from `ggridges`) show kernel density estimates stacked vertically, with fill colour mapped to body size. `scale` controls how much adjacent ridges overlap (0.9 keeps them mostly separate), and `rel_min_height` trims density tails below 1% of the peak to avoid long, thin artefacts:

```{r}
#| label: fig-bodysize-by-biozone
#| code-fold: true
#| fig-cap: "Body size distribution by biozone"
#| fig-height: 5

biozone_plot_data <- occurrences_in_habitat |>
  st_drop_geometry() |>
  filter(!is.na(biozone), !is.na(max_length_mm)) |>
  group_by(biozone) |>
  filter(n() >= 10) |>
  ungroup()

ggplot(
  biozone_plot_data,
  aes(x = max_length_mm, y = biozone)
) +
  geom_density_ridges_gradient(
    aes(fill = after_stat(x)),
    scale = 0.9,
    rel_min_height = 0.01
  ) +
  scale_x_log10() +
  scale_fill_viridis_c(
    name = "Body size (mm)",
    option = "plasma",
    trans = "log10"
  ) +
  labs(
    x = "Maximum body length (mm, log scale)",
    y = "Biozone"
  ) +
  theme_minimal()
```

The deeper biozones (lower bathyal, abyssal) are dominated by smaller-bodied species, while the shallower zones (infralittoral, circalittoral) show broader body size distributions extending to larger species. This pattern is consistent with the shift in taxonomic composition and environmental conditions across depth zones.

We can break this down further by looking at taxonomic class within each biozone, using the same violin + boxplot approach as @fig-bodysize-by-class:

```{r}
#| label: fig-bodysize-class-biozone
#| code-fold: true
#| fig-cap: "Body size distributions by taxonomic class within each biozone (classes with fewer than 5 species excluded)"
#| fig-height: 8
#| fig-width: 10

class_biozone_data <- occurrences_in_habitat |>
  st_drop_geometry() |>
  filter(!is.na(biozone), !is.na(max_length_mm), !is.na(class)) |>
  group_by(biozone) |>
  filter(n() >= 10) |>
  ungroup()

# Keep only classes with >= 5 species across all biozones
class_keep <- class_biozone_data |>
  distinct(aphiaid, class) |>
  count(class) |>
  filter(n >= 5)

class_biozone_data <- class_biozone_data |>
  semi_join(class_keep, by = "class") |>
  mutate(class = reorder(class, max_length_mm, FUN = median))

ggplot(class_biozone_data, aes(x = class, y = max_length_mm)) +
  geom_violin(fill = "#0077BE", alpha = 0.5, colour = "white") +
  geom_boxplot(width = 0.15, outlier.size = 0.5) +
  scale_y_log10() +
  coord_flip() +
  facet_wrap(~biozone) +
  labs(
    x = NULL,
    y = "Maximum body length (mm, log scale)"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

## Interactive Map

Let's create a map showing species occurrences coloured by body size class, overlaid on seabed habitats. We first bin continuous body lengths into discrete size classes using `cut()`:

```{r}
#| label: create-bodysize-classes
# Create body size classes for visualisation
occurrences_for_map <- occurrences_in_habitat |>
  filter(!is.na(max_length_mm)) |>
  mutate(
    size_class = cut(
      max_length_mm,
      breaks = c(0, 10, 50, 200, 1000, Inf),
      labels = c("<10mm", "10-50mm", "50-200mm", "200-1000mm", ">1000mm"),
      right = FALSE
    )
  )
```

The `tmap` package can produce interactive (Leaflet) maps with `tmap_mode("view")`. Here we layer habitat polygons under occurrence points, with pop-ups showing species name and traits:

```{r}
#| label: fig-interactive-map
#| eval: false

tmap_mode("view")

tm_shape(seabed_habitats) +
  tm_polygons(
    fill = "eunis2019c",
    fill.scale = tm_scale_categorical(values = "Set3"),
    fill.legend = tm_legend(title = "EUNIS Habitat"),
    fill_alpha = 0.5,
    col_alpha = 0
  ) +
  tm_shape(occurrences_for_map) +
  tm_symbols(
    fill = "size_class",
    fill.scale = tm_scale_categorical(values = "viridis"),
    fill.legend = tm_legend(title = "Body size"),
    size = 0.1,
    fill_alpha = 0.7,
    popup.vars = c(
      "Species" = "scientificname",
      "Body length (mm)" = "max_length_mm",
      "Habitat" = "EUNIS2019C",
      "Biozone" = "biozone"
    )
  )
```

```{r}
#| label: fig-interactive-map-render
#| echo: false
#| fig-cap: "Interactive map of benthic species occurrences coloured by body size class, overlaid on seabed habitat polygons. Click on features for details."

tmap_mode("view")

interactive_map <- tm_shape(seabed_habitats) +
  tm_polygons(
    fill = "eunis2019c",
    fill.scale = tm_scale_categorical(values = "Set3"),
    fill.legend = tm_legend(title = "EUNIS Habitat"),
    fill_alpha = 0.5,
    col_alpha = 0
  ) +
  tm_shape(occurrences_for_map) +
  tm_symbols(
    fill = "size_class",
    fill.scale = tm_scale_categorical(values = "viridis"),
    fill.legend = tm_legend(title = "Body size"),
    size = 0.1,
    fill_alpha = 0.7,
    popup.vars = c(
      "Species" = "scientificname",
      "Body length (mm)" = "max_length_mm",
      "Habitat" = "EUNIS2019C",
      "Biozone" = "biozone"
    )
  )

# Save to standalone HTML to avoid pandoc hanging on large htmlwidgets
tmap_save(interactive_map, get_output_path("biozone_bodysize_map.html", "tutorials"))

knitr::include_url("biozone_bodysize_map.html", height = "600px")
```

## Discussion

### Key Findings

This analysis demonstrates the integration of multiple EU marine data infrastructures to characterise benthic biozones in the Gulf of Lion:

1.  **Habitat composition varies across biozones**: Shallow zones are dominated by sandy and coarse substrates, while deeper zones shift towards muddy sediments
2.  **Environmental gradients structure biozones**: Depth and temperature covary with biozone, confirming that biozones capture meaningful environmental variation
3.  **Body size shifts with depth**: Shallower biozones tend to support species with larger body sizes, with a shift towards smaller-bodied fauna in deeper zones
4.  **Data integration**: Linking EMODnet occurrence and habitat data with Copernicus environmental layers and external trait databases enables multi-faceted ecological characterisation

### Ecological Interpretation

Biozones represent the vertical zonation of the seabed, reflecting gradients in light, pressure, temperature, and food supply. Our results are consistent with several ecological expectations:

-   **Depth zonation**: The infralittoral and circalittoral zones, receiving more light and organic input, support diverse habitat types and substrates
-   **Body size patterns**: The tendency for smaller body sizes at depth may reflect reduced food availability in deeper zones, consistent with Bergmann's rule inversions observed in some marine taxa
-   **Substrate control**: Substrate composition changes with depth and hydrodynamic regime, shaping habitat availability

However, these interpretations require caution:

1.  **Sampling bias**: Occurrence records may not represent true community composition, and shallow zones are typically better sampled
2.  **Taxonomic coverage**: MOBS trait data is incomplete, with smaller and less-studied taxa underrepresented
3.  **Spatial resolution**: Environmental data resolution may not capture fine-scale habitat heterogeneity

### Limitations

-   **Trait coverage**: Not all species in the occurrence data have body size records in MOBS, so body size patterns reflect a subset of the community
-   **Temporal mismatch**: Occurrence records span different time periods than environmental data
-   **Single region**: Patterns observed in the Gulf of Lion may not generalise to other Mediterranean basins with different oceanographic settings

### Extensions

This workflow could be extended to:

-   Compare biozone characterisation across regions (e.g., Gulf of Lion vs Aegean Sea)
-   Include additional traits from MOBS (feeding mode, mobility, reproduction strategy) for richer functional characterisation
-   Analyse temporal trends in biozone composition
-   Use the biozone framework to assess habitat condition or change over time

## What You've Learned

This tutorial demonstrated advanced multi-source data integration:

-   **EMODnet WFS**: Accessing seabed habitat polygons and species occurrence records
-   **EMODnet WCS**: Retrieving bathymetry rasters for zonal statistics
-   **Copernicus Marine Service**: Downloading bottom temperature data
-   **External databases**: Linking to body size trait data (MOBS) via taxonomic identifiers
-   **Spatial operations**: Zonal statistics, CRS reprojection, and spatial joins across data types
-   **Biozone characterisation**: Using biozones as an organising framework to synthesise habitat, environmental, and trait data

These skills enable complex marine ecological analyses that draw on the full breadth of available EU marine data.

## Further Resources

-   [EMODnet Seabed Habitats Portal](https://emodnet.ec.europa.eu/en/seabed-habitats): Habitat maps
-   [EMODnet Biology Portal](https://emodnet.ec.europa.eu/en/biology): Species occurrence data
-   [Copernicus Marine Service](https://marine.copernicus.eu/): Oceanographic data
-   [World Register of Marine Species (WoRMS)](https://www.marinespecies.org/): Taxonomic backbone
-   [MOBS Database](https://github.com/crmcclain/MOBS_OPEN): Marine body size data

## References
